---
title: "D2 MVA"
output: html_notebook
editor_options: 
  chunk_output_type: console
---
 
 Open dataset

```{r}
library("rstudioapi")
library('ggplot2')
script_dir <- getSourceEditorContext()$path 
setwd(dirname(script_dir))
dd <- read.table("Playstore-reduced.csv",header=T, sep=",");
#Some code for temporal
library(chron)
lct <- Sys.getlocale("LC_TIME"); Sys.setlocale("LC_TIME", "C")
```

```{r}
class(dd)
dim(dd)
n<-dim(dd)[1]
n
K<-dim(dd)[2]
K
names(dd)
str(dd)
summary(dd)
```


Numerical variables: Rating (4), Rating.Count (5), Minimum.Installs (7), Maximum.Installs (8), Price (10)

Binary variables: Free (9),Ad.Supported (21), In.App.Purchases (22), Editors.Choice (23)

Qualitative variables: App.Name (1), App.Id (2), Category (3), Installs (6), Currency (11) ,Size (12), Minimum.Android (13), Developer.Id (14), Developer.Website (15), Developer.Email (16), Content.Rating (19), Privacy.Policy (20)

Temporal: Released (17), Last.Updated (18), Scraped.Time (24) 



Feature selection: filtering the uninteresting variables
```{r}
#Removing irrevalent columns for our analisis

head(summary(dd$App.Id), n = 6)

head(summary(dd$Developer.Id), n = 6)

head(summary(dd$Developer.Website), n = 6)

head(summary(dd$Developer.Email), n = 6)

head(summary(dd$Installs), n = 6)

head(summary(dd$Privacy.Policy), n = 6)

drop <- c("App.Id", "Developer.Id", "Developer.Website", "Developer.Email", "Installs", "Privacy.Policy")
dd = dd[,!(names(dd) %in% drop)]

```


Feature extraction: deriving new variables(DaysLastUpdate,ReleasedDays,AppNameLen)

```{r}
#Released
dd$Released <- as.Date(dd$Released,format='%b %d, %Y')
summary(dd$Released)

#Last.Updated
dd$Last.Updated <- as.Date(dd$Last.Updated,format='%b %d, %Y')
summary(dd$Last.Updated)

#Scraped.Time
dtparts = t(as.data.frame(strsplit(dd$Scraped.Time,' ')))
row.names(dtparts) = NULL
dd$Scraped.Time = chron(dates=dtparts[,1],format='y-m-d')
summary(dd$Scraped.Time)

dd$DaysLastUpdate <- as.numeric(difftime(dd$Scraped.Time,dd$Last.Updated,units="days"))
dd$ReleasedDays <- as.numeric(difftime(dd$Scraped.Time,dd$Released,units="days"))

drop <- c("Released", "Last.Updated", "Scraped.Time")
dd<-  dd[,!(names(dd) %in% drop)]


library(stringr)
appNameLen<-str_length(dd$App.Name)
summary(appNameLen)
dd$AppNameLen <- str_length(dd$App.Name)
drop <- c("App.Name")
dd<-  dd[,!(names(dd) %in% drop)]


```


Transformations;
-Recoding (numeric -> categorical)
-Quantifying a nominal variable (categ. -> numeric))

```{r}
#Size(categorical -> numeric)
dd$Size[dd$Size == "Varies with device"] <- NA
size_kb <- grepl("k", dd$Size)

dd$Size <- gsub("M", "", dd$Size)
dd$Size <- gsub("k", "", dd$Size)
dd$Size <- gsub(",", "", dd$Size)

dd$Size[size_kb] = as.character(as.numeric(dd$Size[size_kb])/1000.0)
dd$Size <- as.numeric(dd$Size)
summary(dd$Size)

```


```{r}
#We set qualitative variables as factors
names(dd)
#Ordinal: Minimum.Android (13), Content.Rating (19)
dd$Minimum.Android<-as.factor(dd$Minimum.Android)
dd$Content.Rating<-as.factor(dd$Content.Rating)

summary(dd$Minimum.Android)
summary(dd$Content.Rating)

#Binary variables: Free (6),Ad.Supported (12), In.App.Purchases (13), Editors.Choice (14)
for(i in c(6,12,13,14)) {
  dd[,i]<-as.factor(dd[,i])
}

summary(dd$Free)

summary(dd$Ad.Supported)

summary(dd$In.App.Purchases)

summary(dd$Editors.Choice)


```


Reducing qualitative variables modalities:
Category (3),Currency (11) , Minimum.Android (13), Content.Rating (19)

```{r}
#Category
#Modalities: Educational, Lifestyle,SocialMedia, Productivity, Entertainment, Game

CatEducational <- c("Books & Reference","Education","Educational","Libraries & Demo","Maps & Navigation","Medical","Parenting","Art & Design","Business","Finance")

CatLifestyle <- c("Food & Drink","Health & Fitness","House & Home","Lifestyle","Personalization","Travel & Local","Shopping","Dating","Photography","Sports","Beauty","News & Magazines","Weather")

CatSocialMedia <- c("Communication","Social","Word")

CatProductivity <- c("Productivity","Tools","Board")

CatEntertainment <- c("Comics","Entertainment","Events","Music","Music & Audio","Video Players & Editors","Auto & Vehicles")

CatGame <- c("Action","Adventure","Arcade","Card","Casino","Casual","Puzzle","Racing","Role Playing","Simulation","Strategy","Trivia")

dd$Category[dd$Category %in% CatEducational] <- "Educational"
dd$Category[dd$Category %in% CatLifestyle] <- "Lifestyle"
dd$Category[dd$Category %in% CatSocialMedia] <- "SocialMedia"
dd$Category[dd$Category %in% CatProductivity] <- "Productivity"
dd$Category[dd$Category %in% CatEntertainment] <- "Entertainment"
dd$Category[dd$Category %in% CatGame] <- "Game"

dd$Category<-as.factor(dd$Category)
summary(dd$Category)


#Minimum.Android

lebAndroid <- as.factor(c("MinimumAndroidUnknown","1.0","1.0","1.0","1.0","2.0","2.0","2.0","2.0","2.0","2.0","2.0","3.0","3.0","3.0","4.0","4.0","4.0","4.0","4.0","4.0","4.0","4.0","4.0","4.0","5.0","5.0","6.0","7.0","7.0","8.0","MinimumAndroidUnknown"))
levels(dd$Minimum.Android) <- lebAndroid
summary(dd$Minimum.Android)
barplot(table(dd$Minimum.Android))


#Currency
dd$Currency <- as.factor(dd$Currency)
levCurrency <- levels(dd$Currency)
levCurrency[levCurrency == ""] = "CurrencyUnknown"
levCurrency[levCurrency == "XXX"] = "CurrencyUnknown"
levels(dd$Currency) <- levCurrency
summary(dd$Currency)

#Content.Rating
summary(dd$Content.Rating)


```


Numerical variables: Rating (4), Rating.Count (5), Minimum.Installs (7), Maximum.Installs (8), Price (10)

```{r}

#Rating
summary(dd$Rating)
boxplot(dd$Rating)
hist(dd$Rating,50)

HighRating_RatingCount <- dd$Rating.Count[dd$Rating == 5]
hist(log10(HighRating_RatingCount),10)
summary(HighRating_RatingCount)
summary(dd$Rating.Count)

plot(dd$Rating,dd$Rating.Count)

#Rating.Count
summary(dd$Rating.Count)
boxplot(dd$Rating.Count)
hist(dd$Rating.Count,100)

log_RatingCount <-log10(dd$Rating.Count)
summary(log_RatingCount)
boxplot(log_RatingCount)
hist(log_RatingCount,100)

head(sort(dd$Rating.Count,decreasing = TRUE))
dd$Rating.Count[dd$Rating.Count > 2000000] <- NA

#Minimum.Installs
summary(dd$Minimum.Installs)
boxplot(dd$Minimum.Installs)
hist(dd$Minimum.Installs,100)

log_MinimumInstalls <-log10(dd$Minimum.Installs + 1)
summary(log_MinimumInstalls)
boxplot(log_MinimumInstalls)
hist(log_MinimumInstalls,100)

#Maximum.Installs
summary(dd$Maximum.Installs)
boxplot(dd$Maximum.Installs)
hist(dd$Maximum.Installs,100)

log_MaximumInstalls <-log10(dd$Maximum.Installs + 1)
summary(log_MaximumInstalls)
boxplot(log_MaximumInstalls)
hist(log_MaximumInstalls,100)

head(sort(dd$Maximum.Installs,decreasing = TRUE))
dd$Maximum.Installs[dd$Maximum.Installs > 200000000] <- NA

#Price
summary(dd$Price)
boxplot(dd$Price)
hist(dd$Price,100)

Price_no_0 <- dd$Price[dd$Price > 0]
summary(Price_no_0)
boxplot(Price_no_0)
hist(Price_no_0,20)

log_Price <-log10(dd$Price + 1)
summary(log_Price)
boxplot(log_Price)
hist(log_Price,100)

log_Price_no_0 <-log10(Price_no_0 + 1)
summary(log_Price_no_0)
boxplot(log_Price_no_0)
hist(log_Price_no_0,10)


#DaysLastUpdate
summary(dd$DaysLastUpdate)
boxplot(dd$DaysLastUpdate)
hist(dd$DaysLastUpdate,100)

log_DaysLastUpdate <-log10(dd$DaysLastUpdate + 1)
summary(log_DaysLastUpdate)
boxplot(log_DaysLastUpdate)
hist(log_DaysLastUpdate,20)

```

Removing duplicated rows
```{r}
#we first check how many duplicted there are
dim(dd[!duplicated(dd),])
#Removing duplicated rows
dd <- dd[!duplicated(dd),]
#checking dataset size after deduplication
dim(dd[!duplicated(dd),])
#there is no duplicated column
```



Comparison between free and payment Apps 
```{r}
pay <- dd[dd$Price > 0,]

free <- dd[dd$Price == 0,]

ks.test(pay$Rating,free$Rating)$p.value

ks.test(pay$Rating.Count,free$Rating.Count)$p.value
ks.test(pay$Minimum.Installs,free$Minimum.Installs)$p.value
ks.test(pay$Maximum.Installs,free$Maximum.Installs)$p.value
ks.test(pay$DaysLastUpdate,free$DaysLastUpdate)$p.value
ks.test(pay$Size,free$Size)$p.value

dim(pay)
dim(free)

```


```{r}
#according to the test result, there is difference between free and payment apps
#we decide to analyze only free apps
dd <- dd[dd$Free =='True',]
#deleting irrevalent columns 
drop <- c("Currency", "Free", "Price", "Editors.Choice")
dd<-  dd[,!(names(dd) %in% drop)]
```


```{r}
dd$Installs <- dd$Maximum.Installs
drop <- c("Maximum.Installs", "Minimum.Installs")
dd<-  dd[,!(names(dd) %in% drop)]
```


The little test
```{r}
library('finalfit')
dd_new=dd
#rename 'unknown' factor level
levels(dd_new$Minimum.Android)[levels(dd_new$Minimum.Android)=='MinimumAndroidUnknown'] <- NA
summary(dd_new$Minimum.Android)

dd_new %>% 
  ff_glimpse()



```

Missing data type
```{r}
library(UpSetR)

dependent<-c('Rating','Rating.Count','Size','Minimum.Android','ReleasedDays')
explanatory<- c('Category','Content.Rating','Ad.Supported','In.App.Purchases','DaysLastUpdate','AppNameLen','Installs')

dd_new %>% 
  missing_pattern(dependent,explanatory)

gg_miss_upset(dd_new)

newdata1 <- dd_new[order(dd_new$Category),]
n1<-gg_miss_case_cumsum(newdata1)

newdata2 <- dd_new[order(dd_new$Rating),]
n2<-gg_miss_case_cumsum(newdata2)

newdata3 <- dd_new[order(dd_new$Rating.Count),]
n3<-gg_miss_case_cumsum(newdata3)

newdata4 <- dd_new[order(dd_new$Size),]
n4<-gg_miss_case_cumsum(newdata4)

newdata5 <- dd_new[order(dd_new$Minimum.Android),]
n5<-gg_miss_case_cumsum(newdata5)

newdata6 <- dd_new[order(dd_new$Content.Rating),]
n6<-gg_miss_case_cumsum(newdata6)

newdata7 <- dd_new[order(dd_new$Ad.Supported),]
n7<-gg_miss_case_cumsum(newdata7)

newdata8 <- dd_new[order(dd_new$In.App.Purchases),]
n8<-gg_miss_case_cumsum(newdata8)

newdata9 <- dd_new[order(dd_new$DaysLastUpdate),]
n9<-gg_miss_case_cumsum(newdata9)

newdata10 <- dd_new[order(dd_new$ReleasedDays),]
n10<-gg_miss_case_cumsum(newdata10)

newdata11 <- dd_new[order(dd_new$AppNameLen),]
n11<-gg_miss_case_cumsum(newdata11)

newdata12 <- dd_new[order(dd_new$Installs),]
n12<-gg_miss_case_cumsum(newdata12)

ggpubr::ggarrange(n1,n2,n3,n4,n5,n6,n7,n8,n9,n10,n11,n12)

library("naniar")
numericalVariablesIndex <- which(sapply(newdata12,is.numeric))
numericalVariablesIndex
mcar_test(newdata12[,numericalVariablesIndex])

```

Rating
```{r}
dd_new %>% 
  missing_compare(dependent[1],explanatory)

gr<-ggplot(dd_new,
       aes(x = Rating,
           y = Installs)) +
  geom_miss_point()

```

Rating.Count
```{r}
dd_new %>% 
  missing_compare(dependent[2],explanatory)

gc<-ggplot(dd_new,
       aes(x = Rating.Count,
           y = Installs)) +
  geom_miss_point()
```

Size
```{r}
dd_new %>% 
  missing_compare(dependent[3],explanatory)

gs<-ggplot(dd_new,
       aes(x = Size,
           y = Installs)) +
  geom_miss_point()
```

Minimum.Android
```{r}
dd_new %>% 
  missing_compare(dependent[4],explanatory)

gm<-ggplot(dd_new,
       aes(x = Minimum.Android,
           y = Installs)) +
  geom_miss_point()
```

ReleasedDays
```{r}
dd_new %>% 
  missing_compare(dependent[5],explanatory)

gd<-ggplot(dd_new,
       aes(x = ReleasedDays,
           y = Installs)) +
geom_miss_point()

ggpubr::ggarrange(gr,gc,gs,gm,gd)
```

Numerical variables imputation
```{r}

dd_imputation<-dd

#imputation of numeric missing values
numericalVariablesIndex <- which(sapply(dd,is.numeric))
numericalVariablesIndex

library("mice")
dd$Installs<-log(dd$Installs+1)
dd$Size<-log(dd$Size+1)
dd$Rating.Count<-log(dd$Rating.Count+1)
numericas<-dd[,numericalVariablesIndex]

res.mice <- mice(numericas,m=10)
str(res.mice)
res.mice$data # Original dataset
completed<-complete(res.mice) # Data frame containing completed data set

summary(completed)


#completed$Installs<-exp(completed$Installs)-1

##correlation map of numerical variables
Correlations <- cor(completed)
print(Correlations)
corrplot::corrplot(Correlations) 
corrplot::corrplot(Correlations, method = "number",number.cex = 0.75) 

dd[,numericalVariablesIndex]<-completed
```


Comparison between populations before and after the imputation of missing data
```{r}
par(mfrow=c(1,2))
hist(numericas[,1],main='Rating')
hist(completed$Rating,main='Rating')

par(mfrow=c(1,2))
hist(numericas[,2],main='Rating.Count')
hist(completed$Rating.Count,main='Rating.Count')

par(mfrow=c(1,2))
hist(numericas[,3],main='Size')
hist(completed$Size,main='Size')

par(mfrow=c(1,2))
hist(numericas[,4],main='DaysLastUpdate')
hist(completed$DaysLastUpdate,main='DaysLastUpdate')

par(mfrow=c(1,2))
hist(numericas[,5],main='ReleasedDays')
hist(completed$ReleasedDays,main='ReleasedDays')

par(mfrow=c(1,1))

```

Categorical variables imputation
```{r}
categoricalVariables <- which(sapply(dd,is.factor))

categoricalVariables

library(missMDA)
#res.mca<-imputeMCA(dd_new[,categoricalVariables])
res.mice<-mice(dd_new[,categoricalVariables],m=10)
completed<-complete(res.mice) 
summary(completed)

par(mfrow=c(1,2))
barplot(prop.table(table(dd[,5])),ylim=c(0,0.9),main="Android_Before")
barplot(prop.table(table(completed[,2])),ylim=c(0,0.9),main="Android_After")
par(mfrow=c(1,1))

dd[,categoricalVariables]<-completed
summary(dd)
```

Filtering by rating count>20
```{r}
dd1 <- dd[dd$Rating.Count < 20 | is.na(dd$Rating.Count),]
ddout <- dd1[,c(1,2,3,4,9,10,11,12)]

dd$Rating.Count<-exp(dd$Rating.Count)-1
summary(dd$Rating.Count)
high <- dd[dd$Rating.Count >= 20,]
low <- dd[dd$Rating.Count < 20,]

ks.test(high$Rating,low$Rating)$p.value
ks.test(high$DaysLastUpdate,low$DaysLastUpdate)$p.value
ks.test(high$Size,low$Size)$p.value

hist(low$Rating,50)
hist(high$Rating,50)
#we keep only rows with rating count greater or equal than 20 
dd <- dd[dd$Rating.Count >= 20 | is.na(dd$Rating.Count),]
dd$Rating.Count<-log(dd$Rating.Count+1)
```

Multivariate outlier detection and treatment
```{r}
X=dd[,numericalVariablesIndex]
X
mdi = mahalanobis(X,center=apply(X,2,mean),cov=var(X))
plot(density(mdi))
cutoff <- qchisq(p = 0.99 , ncol(X))
## Display observation whose distance greater than cutoff value
X[mdi>cutoff,]
#Z<-data.frame(X[mdi>cutoff,],dd[mdi>cutoff,1]
outliers<-data.frame(X[mdi>cutoff,],dd[mdi>cutoff,1])

names(outliers)[names(outliers) == 'dd.mdi...cutoff..1.'] <- 'Category'

doutliers <- rbind(ddout, outliers)
write.csv(doutliers, file.path(getwd(), "Playstore-outliers.csv"), row.names = FALSE)

dd<-dd[!(rownames(dd) %in% rownames(outliers)),]
summary(dd)

write.csv(dd, file.path(getwd(), "Playstore-preprocessed.csv"), row.names = FALSE)
```





