---
title: "Task1"
output: html_document
date: '2023-11-09'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
data <- read.table("TSICHR22RAW.raw", header=TRUE)
genetic_data <- data[, 7:ncol(data)]
```

# 1. How many variants are there in this database? What percentage of the data is missing?

There are 20649 variants.
The percentage of data missing is 0.20.
```{r}
# Counting the number of variants
num_variants <- ncol(genetic_data)
percentage_missing <- mean(is.na(genetic_data)) * 100
```

## 2. Calculate the percentage of monomorphic variants. Exclude all monomorphics from the database for all posterior computations of the practical. How many variants do remain in your database?

A monomorphic variant is one where all individuals have the same genotype. In this case, this would mean all values in a column are the same.
```{r}
is_monomorphic <- apply(genetic_data, 2, function(x) length(unique(na.omit(x))) == 1)
percentage_monomorphic <- sum(is_monomorphic) / length(is_monomorphic) * 100

# Removing monomorphic variants
polymorphic_data <- genetic_data[, !is_monomorphic]
remaining_variants <- ncol(polymorphic_data)

cat("\nRemaining variants:",remaining_variants, "\n")
```


## 3. Report the genotype counts and the minor allele count of polymorphism rs8138488 C, and calculate the MAF of this variant.
```{r}

# Extract the genotype data for rs8138488
rs8138488_genotypes <- na.omit(polymorphic_data$rs8138488_C)

# Calculate the genotype counts for AA, AB, and BB
genotype_counts <- table(rs8138488_genotypes)


aa_count <- genotype_counts["0"]
ab_count <- genotype_counts["1"]
bb_count <- genotype_counts["2"]

total_alleles <- 2 * (aa_count + ab_count + bb_count)
pA <- (2 * aa_count + ab_count) / total_alleles  # frequency of allele A
pB <- (2 * bb_count + ab_count) / total_alleles  # frequency of allele B

# Calculate MAF
maf <- min(pA, pB)

# Output the results
cat("Genotype counts for rs8138488 (AA AB BB):", genotype_counts, "\n")
cat("Minor allele count for rs8138488:", maf * total_alleles, "\n")
cat("MAF for rs8138488:", maf, "\n")


```

## 4.Compute the minor allele frequencies (MAF) for all markers, and make a histogram of it. Does the MAF follow a uniform distribution? What percentage of the markers have a MAF below 0.05? And below 0.01? Can you explain the observed pattern?
```{r}

calculate_maf <- function(genotype) {
  aa_count <- sum(genotype == 0)
  ab_count <- sum(genotype == 1, na.rm = TRUE)
  bb_count <- sum(genotype == 2, na.rm = TRUE)
  
  # Calculate allele frequencies
  total_alleles <- 2 * (aa_count + ab_count + bb_count)
  pA <- (2 * aa_count + ab_count) / total_alleles  # frequency of allele A
  pB <- (2 * bb_count + ab_count) / total_alleles  # frequency of allele B
  
  # Calculate MAF
  return(min(pA, pB))
}


maf_all <- apply(polymorphic_data, 2, calculate_maf)

hist(maf_all, main = "Histogram of Minor Allele Frequencies (MAF)", 
     xlab = "MAF", ylab = "Frequency", col = "blue")


# Calculate the percentage of markers with MAF below 0.05 and 0.01
percent_maf_below_005 <- sum(maf_all < 0.05) / length(maf_all) * 100
percent_maf_below_001 <- sum(maf_all < 0.01) / length(maf_all) * 100

# Output the results
cat("Percentage of markers with MAF below 0.05:", percent_maf_below_005, "%\n")
cat("Percentage of markers with MAF below 0.01:", percent_maf_below_001, "%\n")


```

## 5. Calculate the observed heterozygosity H0, and make a histogram of it. What is, theo- retically, the range of variation of this statistic?
```{r}

# Calculate observed heterozygosity (Ho) for each marker, which is the frequency of the AB genotype
calculate_ho <- function(genotype) {
  ab_count <- sum(genotype == 1, na.rm = TRUE)  # Count the number of AB genotypes
  total_genotypes <- sum(!is.na(genotype))       # Count the total number of non-missing genotypes
  ho <- ab_count / total_genotypes               # Calculate Ho as the proportion of AB genotypes
  return(ho)
}

# Apply the function to all markers (columns) in the genetic data
ho_all <- apply(polymorphic_data, 2, calculate_ho)

hist(ho_all, breaks = 100, main = "Histogram of Observed Heterozygosity (Ho)", 
     xlab = "Ho", ylab = "Frequency", col = "green")

# The theoretical range of variation for Ho is from 0 to 1, where 0 means no heterozygosity
# and 1 means complete heterozygosity.

summary_ho <- summary(ho_all)
print(summary_ho)


```

## 6. Compute for each marker its expected heterozygosity (He), where the expected het- erozygosityforabi-allelicmarkerisdefinedas1−Pki=1p2i ,wherep2iisthefrequencyofthe ith allele. Make a histogram of the expected heterozygosity. What is, theoretically, the range of variation of this statistic? What is the average of He for this database?

He is calculated as 1 - sum of the squared allele frequencies.

```{r}
# Calculate expected heterozygosity (He) for each marker
calculate_he <- function(genotype) {
  aa_count <- sum(genotype == 0, na.rm = TRUE)
  ab_count <- sum(genotype == 1, na.rm = TRUE)
  bb_count <- sum(genotype == 2, na.rm = TRUE)
  
  # Calculate allele frequencies
  total_alleles <- 2 * (aa_count + ab_count + bb_count)
  pA <- (2 * aa_count + ab_count) / total_alleles  # frequency of allele A
  pB <- (2 * bb_count + ab_count) / total_alleles  # frequency of allele B
  
  # Calculate expected heterozygosity
  he <- 1 - (pA^2 + pB^2)
  return(he)
}

# Apply the function to all markers (columns) in the genetic data
he_all <- apply(polymorphic_data, 2, calculate_he)

# Create a histogram of He
hist(he_all, breaks = 100, main = "Histogram of Expected Heterozygosity (He)", 
     xlab = "He", ylab = "Frequency", col = "purple")

# The theoretical range of variation for He is from 0 to 0.5 for bi-allelic markers,
# where 0 means no genetic diversity and 0.5 means maximum diversity.

# Calculate the average He for the dataset
average_he <- mean(he_all)

# Output the average He
cat("Average expected heterozygosity (He):", average_he, "\n")

```


# STR dataset

```{r}
install.packages("HardyWeinberg")
library(HardyWeinberg)
data("NistSTRs")
```

## 1. How many individuals and how many STRs contains the database?

```{r}
num_individuals <- nrow(NistSTRs)
num_STRs <- ncol(NistSTRs) / 2  # Since each STR is represented by two columns (alleles)

num_individuals
num_STRs


# Print the results
cat("Number of individuals:", num_individuals, "\n")
cat("Number of STRs:", num_STRs, "\n")
```

## 2. Write a function that determines the number of alleles for a STR. Determine the number of alleles for each STR in the database. Compute basic descriptive statistics of the number of alleles (mean, standard deviation, median, minimum, maximum)

We iterate over the STRs, treating each STR as a pair of columns. The number of STRs is half the total number of columns (since each STR is represented by two columns).

For each STR, we extract its corresponding pair of columns across all individuals and count the unique values (alleles)

```{r}
# Custom function to determine the number of alleles for a given STR
num_alleles_per_STR <- function(str_index, str_data) {
  alleles_1 <- str_data[, (2 * str_index - 1)]
  alleles_2 <- str_data[, (2 * str_index)]
  
  # Combine alleles from both columns and sort them
  all_alleles <- sort(c(alleles_1, alleles_2))
  
  # Count the unique alleles
  num_unique_alleles <- length(unique(all_alleles))
  
  return(num_unique_alleles)
}

# Number of STRs (columns in the dataset, considering two columns per STR)
num_STRs <- ncol(NistSTRs) / 2

# Determine the number of alleles for each STR in the database
alleles_per_STR <- sapply(1:num_STRs, function(i) num_alleles_per_STR(i, NistSTRs))

# Descriptive statistics
mean_alleles <- mean(alleles_per_STR)
sd_alleles <- sd(alleles_per_STR)
median_alleles <- median(alleles_per_STR)
min_alleles <- min(alleles_per_STR)
max_alleles <- max(alleles_per_STR)

# Return descriptive statistics
list(mean = mean_alleles, sd = sd_alleles, median = median_alleles, min = min_alleles, max = max_alleles)


```

````{r}
alleles_per_STR

````



## 3.  Make a table with the number of STRs for a given number of alleles and present a barplot of the number STRs in each category. What is the most common number of alleles for an STR?

```{r}
allele_counts <- table(alleles_per_STR)
barplot(allele_counts, main = "Number of STRs per Number of Alleles", xlab = "Number of Alleles", ylab = "Number of STRs")

most_common_alleles <- as.integer(names(allele_counts)[which.max(allele_counts)])

```

## 4. Compute the expected heterozygosity for each STR. Make a histogram of the expected heterozygosity over all STRS. Compute the average expected heterozygosity over all STRs.

Expected heterozygosity can be calculated as $1 - \sum (p_i^2)$ where $p_i$ is the frequency of the $i^{th}$ allele. 


```{r}
expected_heterozygosity <- sapply(1:ncol(NistSTRs)/2, function(i) {
  alleles <- table(c(NistSTRs[, (2*i-1)], NistSTRs[, 2*i]))
  1 - sum((alleles/sum(alleles))^2)
})

hist(expected_heterozygosity, main = "Histogram of Expected Heterozygosity", xlab = "Heterozygosity")
mean_expected_heterozygosity <- mean(expected_heterozygosity)
```



## 5. Calculate also the observed heterozygosity for each STR. Plot observed against ex- pected heterozy- gosity, using all STRs. What do you observe?

Observed heterozygosity is the proportion of individuals that are heterozygous for each STR.

```{r}
observed_heterozygosity <- sapply(1:ncol(NistSTRs)/2, function(i) {
  sum(NistSTRs[, (2*i-1)] != NistSTRs[, 2*i]) / nrow(NistSTRs)
})

plot(expected_heterozygosity, observed_heterozygosity, 
     xlab = "Expected Heterozygosity", ylab = "Observed Heterozygosity", 
     main = "Observed vs Expected Heterozygosity")
```


## 6. Compare, overall, the results you obtained for the SNP database with those you obtained for the STR database. What differences do you observe between these two types of genetic markers?

TODO