---
title: "LabSession_6"
author: "Lidia"
date: "Monday, November 21st, 2022"
output: word_document
editor_options: 
  chunk_output_type: console
---

## Prepare Dataset and load packages
```{r}
# Clear plots
if(!is.null(dev.list())) dev.off()

# Clean workspace
rm(list=ls())

load("Elections_92.RData")
options(contrasts=c("contr.treatment","contr.treatment"))
library(car)
library(MASS)
library(AER)
library(effects)
library(lmtest)
library(FactoMineR)
library(DescTools)
library(ResourceSelection)
library(statmod)
library(cvAUC)
```

# Point 1: Exploratory Data Analysis

```{r}
# Feature Selection
summary(elecc92)
catdes(elecc92[,-1],9)
```

# Point 2: Aggregated vs Disaggregated data for binary response models

## Disaggregated data point of view

```{r}
m1 <- glm(pres~age,data=elecc92,family=binomial)
summary(m1)
residualPlots(m1)
marginalModelPlots(m1)
# Manual diagnostic plot using dfage aggregated dataset
# 
par(mfrow=c(1,1))
elecc92$olog <- log((elecc92$pres+0.05)/(1-elecc92$pres+0.05))
plot(elecc92$age,elecc92$olog,pch=19,col="grey70")
lines(lowess(elecc92$age,elecc92$olog,f=0.5),col="blue",lwd=2)
lines(elecc92$age,m1$linear.predictor,col="red",lwd=2)
```

Aggregated data point of view: dfage

```{r}
summary(dfage)
m1a <-glm(cbind(ypos,yneg)~age,data=dfage,family=binomial)
summary(m1a)

# Goodness of fit 

1-pchisq(m1a$dev,m1a$df.res)  # Aggregated data model
1-pchisq(m1$dev,m1$df.res)

# Interpret age effect in m1
exp(0.013491) # Increment 1 year - Odd scale
exp(0.13491) # Increment 10 years 

residualPlots(m1a)
avPlots(m1a)
```

Let's check the fit using a manual plot and also residual analysis methods available in package car 

```{r}
# Manual diagnostic plot using dfage aggregated dataset
# 
par(mfrow=c(1,1))
dfage$olog <- log((dfage$ypos+0.5)/(dfage$yneg+0.5))
plot(dfage$age,dfage$olog,pch=19,col="grey70")
lines(lowess(dfage$age,dfage$olog,f=0.5),col="blue",lwd=2)
lines(dfage$age,m1a$linear.predictor,col="red",lwd=2)

# library(car)
scatterplot(dfage$olog~dfage$age)

# library(effects)
plot(allEffects(m1a),ask=FALSE)
residualPlots(m1a, layout=c(1, 2))
avPlots(m1a,id=list(labels=row.names(dfage), n=5))
outlierTest(m1a)
influenceIndexPlot(m1a,id=list(n=5),vars=c("Cook", "Student","hat"))
influencePlot(m1a,id=list(method="noteworthy", n=5))
```

## Point 3

```{r}
m2 <- glm(pres~age+I((age-42)^2),data=elecc92,family=binomial)
m3 <- glm(pres~age+I((age-42)^2)+I((age-42)^3),data=elecc92,family=binomial)
summary(m3)
anova(m2,m3,test="Chisq")
anova(m1,m2,test="Chisq")

m2p <- glm(pres~poly(age,2),data=elecc92,family=binomial)
m3p <- glm(pres~poly(age,3),data=elecc92,family=binomial)
summary(m3p)

plot(allEffects(m2p),ask=FALSE)
residualPlots(m2p)
marginalModelPlots( m2p )

summary(m2p)
m2a <-glm(cbind(ypos,yneg)~age+I((age-42)^2),data=dfage,family=binomial)
m2ap <-glm(cbind(ypos,yneg)~poly(age,2),data=dfage,family=binomial)
m3a <-glm(cbind(ypos,yneg)~age+I((age-42)^2)+I((age-42)^3),data=dfage,family=binomial)
summary(m3a)
anova(m2a,m3a,test="Chisq")
anova(m1a,m2a,test="Chis")

AIC(m3,m2,m1)# Min AIC: 2652.6
step(m3,k=log(nrow(elecc92)))

AIC(m3a,m2a,m1a)  # Warning: grouping consistency should be guaranteed
step(m3a)

marginalModelPlots(m2p,id=list(labels=row.names(elecc92), n=5)) 
avPlots(m2p,id=list(labels=row.names(elecc92), n=5))
avPlots(m2ap,id=list(labels=row.names(dfage), n=5))

residualPlots(m2p, layout=c(1, 2))
residualPlots(m2ap, layout=c(1, 2))
outlierTest(m2p)
influenceIndexPlot(m2p,id=list( n=5))
influenceIndexPlot(m2ap,id=list( n=5))

# library(lmtest)
waldtest(m1, m2, test = "Chisq") # Wald Test
waldtest(m2, m3, test = "Chisq") # Wald Test

waldtest(m1a, m2a, test = "Chisq") # Wald Test
waldtest(m2a, m3a, test = "Chisq") # Wald Test
```

## Point 4

```{r}
######
summary(elecc92)

m1c <- glm(pres~c.age,data=elecc92,family=binomial)
summary(m1c)

# Interpret 40 to 59 age group
# 
exp( 0.83446 );100*(exp( 0.83446 ) -1)

residualPlots(m1c , layout=c(1, 2))
marginalModelPlots(m1c,id=list(n=5))
```

## Point 5

```{r}
m1c <- glm(pres~c.age,data=elecc92,family=binomial)
AIC(m1,m2,m3,m1c)# Min AIC: 2652.608 m2
```

## Point 6

```{r}
m4  <-glm(pres~ poly(age,2)+poly(educ,3),data=elecc92,family=binomial)
summary(m4)

residualPlots(m4 , layout=c(2, 2))
marginalModelPlots(m4,id=list(n=5)) 
avPlots(m4,id=list(labels=row.names(elecc92), n=5))
Anova(m4, test="LR")
```

## Point 7

```{r}
m5 <-glm(pres~poly(age,2)+c.edu,data=elecc92,family=binomial)
m6 <-glm(pres~poly(age,2)*c.edu,data=elecc92,family=binomial)

plot(allEffects(m6),ask=FALSE)

anova( m5, m6, test="Chisq")
#library(lmtest)
waldtest(m5,m6,test='Chisq')

marginalModelPlots(m6) 
residualPlots(m6)
```

## Point 8:

```{r}
AIC(m4,m5,m6)
```

## Party Preferences

```{r}
elecc92$party <- as.factor(paste(elecc92$p,elecc92$arty))
summary(as.factor(c(elecc92$p,elecc92$arty)))
summary(elecc92)
m7<- glm(pres ~ poly(age,2)*c.edu+party, family=binomial, data=elecc92)
summary(m7)

# Grouping levels
plot(vote~party,data=elecc92)
partit <- rep("weak",length(elecc92$party))
partit[elecc92$party=="strong rep"|elecc92$party=="strong dem"] <- "strong"
partit[elecc92$party=="ind ind"] <- "ind"
partit <-ordered(partit,levels=c("ind","weak","strong"))
newparty <-partit
elecc92 <- data.frame(elecc92,newparty)

summary(elecc92)

m8<- glm(pres ~ poly(age,2)*c.edu+newparty, family=binomial, data=elecc92)
m9<- glm(pres ~ poly(age,2)*c.edu*newparty, family=binomial, data=elecc92)
summary(m8)
Anova( m8, test="LR")
Anova( m9, test="LR")
anova( m8, m9, test="Chisq")
waldtest(m8,m9, test="Chisq")

anova(m8,m7, test="Chisq")
#waldtest(m8,m7, test="Chisq") # Actually, they are nested
AIC(m8,m7,m9)

plot(allEffects(m8),ask=FALSE)
```

# Point 10: Interest in the elections

```{r}
options(contrasts=c("contr.treatment","contr.treatment"))
m10<- glm(pres ~ poly(age,2)*c.edu+newparty+inter, family=binomial, data=elecc92)
m10i<- glm(pres ~(poly(age,2)*c.edu+newparty)*inter, family=binomial, data=elecc92)

anova( m10, m10i, test="Chisq")
waldtest( m10, m10i, test="Chisq" )
AIC( m8, m10, m10i )

plot(allEffects(m10),ask=F)

# Test de Wald
summary( m10 )
waldtest( m8, m10, test="Chisq")
beta <- as.numeric(m10$coe[9:10]);beta # test de Wald
library(MASS)
V <- vcov(m10)[9:10,9:10] # Var-Cov Matrix Estimates
Wald <- beta%*%solve(V)%*%beta; Wald # Wald Statistic
```

## Point 11: interaCtion interest and new party preferences

```{r}
m11 <- update(m10,~.+inter:newparty)
summary(m11)
anova(m10, m11,test="Chisq")
waldtest(m10, m11,test="Chisq") # Not worth
```

## Point 12: Close results

```{r}
m12 <- update(m10,~.+close)
summary(m12)
anova(m10, m12,test="Chisq")
waldtest(m10, m12,test="Chisq")
```

## Point 13: Satisfaction with candidatures

```{r}
m13 <- update(m10,~.+sat)
summary(m13)
anova(m10,m13, test="Chisq")

Anova( m13, test="LR")
```

## Point 14 - Split data into work and test

```{r}
set.seed(1234)
llwork <- sample(1:nrow(elecc92),round(0.8*nrow(elecc92),dig=0))
dfwork <- elecc92[llwork,]
dftest <- elecc92[-llwork,]

m14<- glm(pres ~ (poly(age, 2) * c.edu + newparty)*(inter+close+sat), family=binomial, data=dfwork)
m15 <- step(m14, k=log(nrow(dfwork)))
m16 <- step(m14)
Anova(m15, test="LR")
Anova(m16, test="LR")

anova(m15, m16, test="Chisq")
summary(m15)

# Diagnostics

marginalModelPlots(m15,id=list(labels=row.names(elecc92),method=abs(cooks.distance(m15)), n=5) )

avPlots(m15,id=list(labels=row.names(elecc92),method=abs(cooks.distance(m15)), n=5) )

crPlots(m15,id=list(labels=row.names(elecc92),method=abs(cooks.distance(m15)), n=5) )

residualPlots(m15, layout=c(3, 2))
outlierTest(m15)
llnrem<-influencePlot(m15,id=list(n=3) );llnrem

llistal<-Boxplot(hatvalues(m15), id.n=10)
abline(h=4*length(coef(m15))/nrow(dfwork))
hatvalues(m15)[llistal]

llistac<-Boxplot(cooks.distance(m15), id=list(labels=rownames(dfwork)))
cooks.distance(m15)[llistac]
elecc92[llistac,]
llrem <- which( rownames(dfwork) %in% c("913","951"));llrem

dfwork <- dfwork[-llrem,]
m15<- glm(pres ~ poly(age, 2) + c.edu + newparty+inter, family=binomial, data=dfwork )
```

# Point 15: Forecasting capability of the final model

```{r}
elecc92.fin <- m15
prob.vot <- elecc92.fin$fit
pres.est <- ifelse(prob.vot<0.5,0,1)
tt <- table(pres.est,dfwork$pres)
sum(diag(table(pres.est,dfwork$pres)))/(dim(dfwork)[1])
precision <- tt[2,2]/(tt[2,2] +tt[2,1] );precision
recall <- tt[2,2]/(tt[2,2] +tt[1,2] );recall
f1 <-2*((precision*recall)/(precision+recall)); f1

# Model na?ve
m0<-glm(pres ~ 1, family=binomial, data=dfwork)
prob.vot0 <- m0$fit
pres.est0 <- ifelse(prob.vot0<0.5,0,1)
tt0 <- table(pres.est0,dfwork$pres)
tt0
table(pres.est0,dfwork$pres)[1,2]/(dim(dfwork)[1])
precision0 <- tt0[1,2]/(tt0[1,2] +tt0[1,1] );precision0
recall0 <- tt0[1,2]/(tt0[1,2] +0 );recall0
f10 <-2*((precision0*recall0)/(precision0+recall0)); f10

library(rms)
model.final <- lrm(pres ~ poly(age, 2) + c.edu + newparty + inter,  data=dfwork)
model.final

library(fmsb)
NagelkerkeR2(m15)
summary(m15)
100*(1-m15$dev/m15$null.dev)

library(DescTools)
PseudoR2(elecc92.fin, which='all') # Not working for grouped data
# Sheather
1 - (elecc92.fin$deviance / elecc92.fin$null.deviance)
# McFadden
1-(as.numeric(logLik(elecc92.fin))/as.numeric(logLik(m0)))

library(ResourceSelection)
hoslem.test(dfwork$pres, fitted(elecc92.fin))

# Estadístic de Hosmer-Lemershow
# seque<-c(seq(0,1,by=0.1));seque  # Original proposal
seque<-quantile(fitted(m15),probs=seq(0,1,by=0.1))
fitgrup<-cut(fitted(m15),breaks=seque); fitgrup
cfitgrup<-table(dfwork$pres,fitgrup);cfitgrup
cmodel1<-tapply(fitted(m15),fitgrup,sum);cmodel1
cmodel0<-tapply(1-fitted(m15),fitgrup,sum);cmodel0
cmodelgrup <- rbind(cmodel0,cmodel1);cmodelgrup
XHL<-sum(((cfitgrup-cmodelgrup)^2)/cmodelgrup);XHL
dfXHL<-8
1-pchisq(XHL,dfXHL)

library("ROCR")
dadesroc<-prediction(predict(m15,type="response"),dfwork$pres)
par(mfrow=c(1,2))
plot(performance(dadesroc,"err"))
plot(performance(dadesroc,"tpr","fpr"))
abline(0,1,lty=2)

library(cvAUC)
AUC(predict(m15,type="response"),dfwork$pres)

```

# Point 15: Forecasting capability of the final model. Test sample

```{r}
# It assumes sample split into dfwork and dftest
# Definition of new party
dftest$party <- as.factor(paste(dftest$p,dftest$arty))
summary(as.factor(c(dftest$p,dftest$arty)))
# Grouping levels
partit <- rep("weak",length(dftest$party))
partit[dftest$party=="strong rep"|dftest$party=="strong dem"] <- "strong"
partit[dftest$party=="ind ind"] <- "ind"
newparty <-factor(partit,levels=c("ind","weak","strong"))
dftest <- data.frame(dftest,newparty)

elecc92.fin <- m15
prob.vot <- predict(elecc92.fin, newdata=dftest, type="response")
pres.est <- ifelse(prob.vot<0.5,0,1)
table(pres.est,dftest$pres)
sum(diag(table(pres.est,dftest$pres)))/dim(dftest)[1]
# Model na?ve
m0<-glm(pres ~ 1, family=binomial, data=dftest)
prob.vot0 <- m0$fit
pres.est0 <- ifelse(prob.vot0<0.5,0,1)
table(pres.est0,dftest$pres)
table(pres.est0,dftest$pres)[1,2]/dim(dftest)[1]

library(ResourceSelection)
hoslem.test(dftest$pres, prob.vot)
```
