---
title: "BSG-MDS practical 6 Statistical Genetics"
author: "Eliya Tiram and Ximena Moure"
date: "12/12/2023, submission deadline 19/12/2023"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("gdsfmt")
#BiocManager::install("SNPRelate")
#install.packages("gdsfmt")
library(genetics)
library(data.table)
library(ggplot2)
library(SNPRelate)
library(gdsfmt)
```


## 1.Load the YRI06.raw file into the R environment. How many individuals and how many SNPs are there in the database? What percentage of the data is missing?

```{r}
# Load data
yri6 <- fread("YRI6.raw")

yri6_snps <- yri6[, 7:ncol(yri6)]

yri6_snps <- as.data.frame(yri6_snps)


# Number of individuals and SNPs
num_individuals <- nrow(yri6_snps)
num_snps <- ncol(yri6) - 6  # Subtracting non-SNP columns

# Calculate missing data percentage
missing_data_percentage <- sum(is.na(yri6[, 7:ncol(yri6)])) / 
  (num_individuals * num_snps) * 100

# Output
cat("Number of individuals:", num_individuals, "\n")
cat("Number of SNPs:", num_snps, "\n")
cat("Percentage of missing data:", missing_data_percentage, "%\n")
```

## 2. Compute, for each pair of individuals (and report the first 5), the mean m of the number of alleles shared and the standard deviation s of the number of alleles shared.

```{r}

# Function to calculate shared alleles
shared_alleles <- function(a, b) {
  2 - abs(a - b)
}

num_individuals <- nrow(yri6)
num_combinations <- choose(num_individuals, 2)
results <- matrix(nrow = num_combinations, ncol = 3)
colnames(results) <- c("Pair", "m", "s")

# Generate combinations of individuals
combos <- combn(num_individuals, 2)
counter <- 1

# Loop through each combination of individuals
for (i in 1:ncol(combos)) {
  ind1 <- combos[1, i]
  ind2 <- combos[2, i]
  
  # Extract SNP data for the pair
  snps1 <- as.integer(yri6[ind1, 6:ncol(yri6)])
  snps2 <- as.integer(yri6[ind2, 6:ncol(yri6)])

  shared <- shared_alleles(snps1, snps2)
  
  
  m <- mean(shared, na.rm = TRUE)
  s <- sd(shared, na.rm = TRUE)
  
  results[counter, ] <- c(paste("Ind", ind1, "- Ind", ind2), m, s)
  counter <- counter + 1
}

results_df <- as.data.frame(results)
results_df$m <- as.numeric(as.character(results_df$m))
results_df$s <- as.numeric(as.character(results_df$s))

head(results_df, 5)

```

## 3.Compute, for each pair of individuals (and report the first 5), the fraction of variants for which the individuals share 0 alleles (p0), and the fraction of variants for which the individuals share 2 alleles (p2). Check if m = 1 − p0 + p2 holds.

```{r}

results_part3 <- data.frame(
  Pair = character(),
  p0 = numeric(),
  p2 = numeric(),
  m = numeric(),
  m_check = logical()
)

for (i in 1:(nrow(yri6) - 1)) {
  for (j in (i + 1):nrow(yri6)) {
    snp_data_i <- unlist(yri6[i, 6:ncol(yri6)])
    snp_data_j <- unlist(yri6[j, 6:ncol(yri6)])
    
    # Shared alleles calculation
    shared <- 2 - abs(snp_data_i - snp_data_j)
    #shared <- mapply(share, snp_data_i, snp_data_j)
    
    
    # Calculate p0, p2, and check if m = 1 - p0 + p2
    p0 <- mean(shared == 0, na.rm = TRUE)
    p2 <- mean(shared == 2, na.rm = TRUE)
    m <- mean(shared, na.rm = TRUE)
    m_check <- abs(m - (1 - p0 + p2)) < 0.0001
    
    # Add the results to the data frame
    results_part3 <- rbind(results_part3, data.frame(
      Pair = paste("Ind", i, "- Ind", j),
      p0 = p0,
      p2 = p2,
      m= m,
      m_check = m_check
    ))
  }
}


str(results_part3)

head(results_part3, 5)


```

## 4. Plot m against s and plot p0 against p2. Comment on the results.

Mean vs. Standard Deviation of Shared Alleles Plot:

- The majority of data points are clustered in a region that indicates a 
relatively high mean of shared alleles (around 1.3) and a low standard deviation
(around 0.55 to 0.65). This clustering suggests that for most individual pairs, 
there is a consistent level of allele sharing.

- There is an outlier group with a higher mean (around 1.5) and very low 
standard deviation (close to 0.5). This could indicate a subset of pairs with a 
higher degree of relatedness, such as parent-offspring pairs, where we would 
expect them to share more alleles on average.

Fraction of Variants Shared 0 Alleles vs. 2 Alleles Plot:

- Most points are clustered towards the lower end of the x-axis (p0) and the 
higher end of the y-axis (p2), suggesting that for the majority of individual 
pairs, there is a low fraction of variants with no shared alleles and a higher 
fraction with both alleles shared. This pattern might indicate some degree of 
relatedness among the individuals.

- There is an outlier group with a high fraction of variants shared 0 alleles 
(p0 around 0.1) and a low fraction of variants shared 2 alleles (p2 around 
0.48). This could represent pairs of individuals that are less related or 
unrelated.


In summary, the data seems to show a general trend of shared genetic similarity 
with some distinct groups that could represent different degrees of relatedness.


```{r}

# Plot m against s
plot_m_vs_s <- ggplot(results_df, aes(x = m, y = s)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Mean vs. Standard Deviation of Shared Alleles",
       x = "Mean of Shared Alleles (m)",
       y = "Standard Deviation of Shared Alleles (s)")

print(plot_m_vs_s)

results_part3$p0 <- as.numeric(as.character(results_part3$p0))
results_part3$p2 <- as.numeric(as.character(results_part3$p2))

# Plot p0 against p2
plot_p0_vs_p2 <- ggplot(results_part3, aes(x = p0, y = p2)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Fraction of Variants Shared 0 Alleles vs. 2 Alleles",
       x = "Fraction of Variants Shared 0 Alleles (p0)",
       y = "Fraction of Variants Shared 2 Alleles (p2)")

print(plot_p0_vs_p2)

```


## 5.Plot m against s and use the pedigree information of the YRI06.raw file to label the data points in the scatterplot. Recall that column 3 and 4 from the YRI06.raw contain information about the family relationship of the participants. Create two labels: one for individuals that have a parent-offspring relationship and another one for unrelated individuals. Comment on the results.

There is a distinct point (or potentially a cluster of points) on the far right,
which has a very high mean number of shared alleles (around 1.5) and a very 
low standard deviation.
They are labeled as a "Parent-Offspring" relationship, which is consistent with 
the expectation that a parent and offspring would share approximately 50% of 
their alleles, leading to a high mean of shared alleles.

The remaining points, labeled as "Unrelated," have lower means (approximately 
between 1.3 and 1.4) and higher standard deviations (approximately between 
0.55 and 0.65).
These points' positions suggest that there is more variability in the number of 
shared alleles among unrelated individuals, as indicated by the higher standard 
deviation.

Overall, the clear separation between the "Parent-Offspring" point and the 
"Unrelated" points indicates that the genetic data effectively differentiates 
between these two types of relationships.

```{r}

is_parent_offspring <- function(ind1, ind2, data) {
  parents_ind1 <- data[IID == ind1, .(PAT, MAT)]
  parents_ind2 <- data[IID == ind2, .(PAT, MAT)]
  
  return((ind1 %in% c(parents_ind2$PAT, parents_ind2$MAT)) | 
         (ind2 %in% c(parents_ind1$PAT, parents_ind1$MAT)))
}

label_matrix <- matrix(NA, nrow = nrow(yri6), ncol = nrow(yri6))
rownames(label_matrix) <- yri6$IID
colnames(label_matrix) <- yri6$IID

for (i in 1:(nrow(yri6) - 1)) {
  for (j in (i + 1):nrow(yri6)) {
    # Check if the pair is parent-offspring
    if (is_parent_offspring(yri6$IID[i], yri6$IID[j], yri6)) {
      label_matrix[i, j] <- "Parent-Offspring"
      label_matrix[j, i] <- "Parent-Offspring"  # Symmetry
    } else {
      label_matrix[i, j] <- "Unrelated"
      label_matrix[j, i] <- "Unrelated"  # Symmetry
    }
  }
}

# Applying the function to label each pair as parent-offspring or unrelated
results_df$Label <- mapply(is_parent_offspring, 
                          as.character(yri6$IID[combos[1,]]), 
                          as.character(yri6$IID[combos[2,]]), 
                          MoreArgs = list(data = yri6))

# Correcting the labels from TRUE/FALSE to Parent-Offspring/Unrelated
results_df$Label <- factor(results_df$Label, 
                           labels = c("Unrelated", "Parent-Offspring"))

# Create the scatter plot
plot_m_vs_s_labeled <- ggplot(results_df, aes(x = m, y = s, color = Label)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Mean vs. Standard Deviation of Shared Alleles",
       x = "Mean Number of Shared Alleles (m)",
       y = "Standard Deviation of Shared Alleles (s)",
       color = "Relationship") +
  scale_color_manual(values = c("Unrelated" = 'red', 
                                "Parent-Offspring" = 'blue'))

# Print the labeled plot
print(plot_m_vs_s_labeled)


```





## 6. Use the package SNPRelate to estimate the IBD probabilities, and plot the probabilities of sharing 0 and 1 IBD alleles (k0 and k1) for all pairs of individuals. Use the pedigree information of the YRI06.raw file to label the data points in the scatterplot (same as before, one colour for parent-offspring relationship and another colour for unrelated individuals).

```{r}


yri6_snps_matrix <- as.matrix(yri6_snps)
sample_ids <- yri6$IID

# Create the GDS file
gds_file <- "genotype_data.gds"
snpgdsCreateGeno(gds_file, yri6_snps_matrix, sample.id = sample_ids, 
                 snp.id = colnames(yri6_snps_matrix), snpfirstdim = FALSE)

genofile <- snpgdsOpen(gds_file)

# Perform LD pruning
snpset <- snpgdsLDpruning(genofile, ld.threshold = 0.2)
snpset.id <- unlist(unname(snpset))

# Estimate IBD probabilities
ibd <- snpgdsIBDMLE(genofile, maf = 0.05, missing.rate = 0.05, 
                    snp.id = snpset.id)

# Select IBD estimates using the 'ibd' method
ibd_selection <- snpgdsIBDSelection(ibd)


snpgdsClose(genofile)

# Assign labels from the labels_matrix to the IBD selection data frame
ibd_selection$label <- mapply(function(id1, id2) label_matrix[id1, id2], 
                              ibd_selection$ID1, ibd_selection$ID2)

# Plot using ggplot2
library(ggplot2)
ggplot(ibd_selection, aes(x = k0, y = k1, color = label)) +
  geom_point() +
  geom_abline(intercept = 1, slope = -1, linetype = "dashed", color = "black") +
  scale_color_manual(values = 
                       c("Unrelated" = 'red', "Parent-Offspring" = 'blue')) +
  labs(title = "IBD Probabilities (k0 vs k1) with Relationships",
       x = "Probability of Sharing 0 IBD Alleles (k0)",
       y = "Probability of Sharing 1 IBD Allele (k1)",
       color = "Relationship")


```

## 7.Do you think the family relationships between all individuals were correctly specified?

From the IBD plot, we observe that:

- The blue points, representing parent-offspring pairs, are closely clustered at 
high values of sharing 1 IBD allele (k1) and low values of sharing 0 IBD 
alleles (k0). This cluster suggests a correct specification of parent-offspring 
relationships because such pairs are expected to share a large proportion of 
their genomes identically by descent.
- The red points, representing unrelated pairs, are spread across the plot with 
a trend that when the probability of sharing 0 IBD alleles (k0) increases, the 
probability of sharing 1 IBD allele (k1) decreases. This is a typical pattern 
for unrelated individuals in a population.

From Part 5, the distinct separation between the "Parent-Offspring" pair and 
the "Unrelated" pairs in the mean vs. standard deviation of shared alleles plot 
suggests that the genetic data is generally consistent with the expected 
relationships. Parent-offspring pairs show a higher mean and lower standard 
deviation, which is consistent with the inheritance patterns.

For Part 4, if the fraction of variants shared 0 alleles vs. 2 alleles plot 
showed a similar consistency and no anomalies, it would further support the 
correct specification of relationships.

Considering the clustering and separation observed in the IBD plot, along with 
consistent results in the mean vs. standard deviation and the fraction of 
variants shared plots, there is a strong indication that family relationships 
have been correctly specified for the most part. The genetic data appears to 
align well with the expected patterns of allele sharing for both 
parent-offspring and unrelated pairs.


