---
title: "Simple Linear Regression_lab - Pre"
author: "Lídia Montero"
date: "2022"
output: word_document
editor_options: 
  chunk_output_type: console
---

# Load Data 

```{r}


# Clear plots
if(!is.null(dev.list())) dev.off()

# Clean workspace
rm(list=ls())

load("Anscombe73raw.RData")

ls()
anscombe

attach(anscombe)
summary(anscombe)

```



Nothing remarkable about suitability of a linear model to address the relationship of Y and X. Linear models is Y=3+0.5X

```{r}
# Set A
par(mfrow=c(1,1))
plot(XA,YA,pch=19)

ma<-lm(YA~XA,data=anscombe)
ls()
summary(ma)
lines(XA,fitted.values(ma))
text(XA,YA,labels=row.names(anscombe),adj=-1)
par(mfrow=c(2,2))
plot(ma)
```

## Set B

A linear model does not address properly the relationship of Y and X, since non linear relation can be seen.
Residual analysis is a validation tool that supports model validation: residuals do not shown a random pattern, a sistematic profile is seen.

```{r}
# Set B
par(mfrow=c(1,1))
plot(XB,YB,pch=19,col="red")
mb<-lm(YB~XB,data=anscombe)
lines(XB,fitted(mb),col="red")
ls()
summary(mb)

par(mfrow=c(2,2))
plot(mb)
mbb<-lm(YB~XB+I(XB^2),data=anscombe)
summary(mbb)

plot(YB~XB,pch=19,col="red")

```

## Set C

A linear model does address the relationship of Y and X, but one observation does not follow the common pattern supported by the rest of the observations.
Residual analysis is a validation tool that supports model validation: residuals do not shown a random pattern, a residual outlier is seen and normal distribution of residuals does not hold.
Observation 3 is influent data, since it affects model parameters. A remarkable change in model parameters is seen whether considering obs. 3, or not.

```{r}
# Set C
par(mfrow=c(1,1))
plot(XC,YC,pch=19,col="darkgreen")
text(XC,YC,label=row.names(anscombe),col="darkgreen",adj=1.5)
mc<-lm(YC~XC,data=anscombe)
lines(XC,fitted(mc),col="darkgreen")
ls()
summary(mc)

par(mfrow=c(2,2))
plot(mc)

library(car)
Boxplot(resid(mc),col="darkgreen")
cooks.distance(mc)
Boxplot(cooks.distance(mc),col="darkgreen")

mcc<-lm(YC~XC,data=anscombe[-3,])
summary(mcc)
df <-anscombe[-3,]
plot(df$XC,df$YC,pch=19,col="darkgreen")
text(df$XC,df$YC,label=row.names(df),col="darkgreen",adj=1.5)
```
## Set D

A linear model does address the relationship of Y and X, but one observation does not follow the common pattern supported by the rest of the observations.
Residual analysis is a validation tool that supports model validation: residuals do not shown a random pattern, a 0 residual is found for 1 observation.
Observation 8 is influent data, since it greatly affects model parameters. Model parameters are totally depending on one observation (obs. 8).


```{r}
# Set D
par(mfrow=c(1,1))
plot(XD,YD,pch=19,col="blue")
text(XD,YD,label=row.names(anscombe),col="blue",adj=1.5)
md<-lm(YD~XD,data=anscombe)
lines(XD,fitted(md),col="blue")
ls()
summary(md)

par(mfrow=c(2,2))
plot(md)

resid(md)
cooks.distance(md)
hatvalues(md)

mdd<-lm(YD~XD,data=anscombe[-8,])
summary(mdd)
mean(YD[-8])
df <-anscombe[-8,]
plot(df$XD,df$YD,pch=19,col="darkgreen")
text(df$XD,df$YD,label=row.names(df),col="darkgreen",adj=1.5)
```

All in all

```{r}
### Prospecció dels jocs de dades A,B,C,D
par(mfrow=c(2,2))
anscombe.lmA <- lm(anscombe$YA ~ anscombe$XA, data=anscombe)
plot(anscombe$XA, anscombe$YA,pch=19,col=1)
lines(anscombe$XA,anscombe.lmA$fitted.values, col=1, lty=3,lwd=2)
text(x=anscombe$XA,y=anscombe$YA,labels=row.names(anscombe),adj=1.1, col=1)

anscombe.lmB <- lm(anscombe$YB ~ anscombe$XB, data=anscombe)
plot(anscombe$XB, anscombe$YB,pch=19,col=2)
lines(anscombe$XB,anscombe.lmB$fitted.values, col=2, lty=3,lwd=2)
text(x=anscombe$XB,y=anscombe$YB,labels=row.names(anscombe),adj=1.1, col=2)

anscombe.lmC <- lm(anscombe$YC ~ anscombe$XC, data=anscombe)
plot(anscombe$XC, anscombe$YC,pch=19,col=3)
lines(anscombe$XC,anscombe.lmC$fitted.values, col=3, lty=3,lwd=2)
text(x=anscombe$XC,y=anscombe$YC,labels=row.names(anscombe),adj=1.1, col=3)

anscombe.lmD <- lm(anscombe$YD ~ anscombe$XD, data=anscombe)
plot(anscombe$XD, anscombe$YD,pch=19,col=4)
lines(anscombe$XD,anscombe.lmD$fitted.values, col=4, lty=3,lwd=2)
text(x=anscombe$XD,y=anscombe$YD,labels=row.names(anscombe),adj=1.1, col=4)


```
