---
title: "LAB Session 7 - Polytomic Outcome"
author: "Lídia Montero"
date: "28 Nov 2022"
output: word_document
editor_options: 
  chunk_output_type: console
---

# Load libraries and set current directory

```{r}
# Clear plots
if(!is.null(dev.list())) dev.off()

# Clean workspace
rm(list=ls())
options(contrasts=c("contr.treatment","contr.treatment"))

library(car)
library(FactoMineR)
library(lmtest)
library(effects)
library(AER)
library(nnet)
library(MNLpred)
library(ggplot2)
library(scales)
```

# Womenlf

```{r}
load("Womenlf.RData")
summary(womenlf)
str(womenlf)
df <- womenlf

Boxplot(womenlf$income, id=list(n=4))
intervals <- c(0,  5 ,10 ,15 ,20 ,25 ,30, 46)
womenlf$gincome <- factor(cut(womenlf$income,breaks=intervals))
table(womenlf$gincome)
cintervals<- tapply(womenlf$income,womenlf$gincome,median)
womenlf$ngincome<-factor(cut(womenlf$income,breaks=intervals))
womenlf$ngincome<-factor(womenlf$ngincome,labels=cintervals)
womenlf$ngincome<-as.numeric(levels(womenlf$ngincome))[womenlf$ngincome]

womenlf$bwork<-ifelse(womenlf$work=="not_work",0,1)
womenlf$bwork<-factor(womenlf$bwork,labels=c("not_work","work"))
```

# Point 2

```{r}
hm1m0 <- glm( bwork ~ 1, family="binomial",data=womenlf)
hm1m1 <- glm( bwork ~ sons, family="binomial",data=womenlf)
hm1m2 <- glm( bwork ~ income, family="binomial",data=womenlf)
hm1m3 <- glm( bwork ~ sons + income, family="binomial",data=womenlf)
hm1m4 <- glm( bwork ~ region, family="binomial",data=womenlf)
hm1m5 <- glm( bwork ~ sons + region, family="binomial",data=womenlf)
hm1m6 <- glm( bwork ~ sons * income, family="binomial",data=womenlf)
hm1m7 <- glm( bwork ~ sons + region + income, family="binomial",data=womenlf)
hm1m8 <- glm( bwork ~ region + sons*income, family="binomial",data=womenlf)
hm1m9 <- glm( bwork ~ region + sons*poly(income,2), family="binomial",data=womenlf)

hm1m1$null.dev - hm1m1$dev
hm1m4$null.dev - hm1m4$dev
hm1m2$null.dev - hm1m2$dev

AIC(hm1m1,hm1m4,hm1m2, hm1m3,hm1m6) # hm1m3 325.7325
step(hm1m9)
anova(hm1m8, hm1m9, test="Chisq")
plot(allEffects(hm1m3))

# Interpreting model parameters
summary(hm1m3)

# Sons effect
coef(hm1m3)
exp(coef(hm1m3))
100*(1-exp(coef(hm1m3))[2])
# Having sons decreases the relative probability of working vs non_working by 79.31% all else being equal
prop.table(table(womenlf$bwork))
# In terms of probability 
prop.table(table(womenlf$bwork))[1]*prop.table(table(womenlf$bwork))[2]*coef(hm1m3)[2]

# Income effect: odds of working decrease for each year by 4.14% and in 0.01 probability units
coef(hm1m3)[3]
exp(coef(hm1m3))[3]
100*(1-exp(coef(hm1m3))[3])
prop.table(table(womenlf$bwork))[1]*prop.table(table(womenlf$bwork))[2]*coef(hm1m3)[3]
```

## Nominal models

```{r}
# Point 3
mm0 <-multinom( work~1, data=womenlf,weight=ones )
mm1 <-multinom( work~sons, data=womenlf,weight=ones )
mm2 <-multinom( work~income, data=womenlf,weight=ones )
mm4 <-multinom( work~region, data=womenlf,weight=ones )
mm3 <-multinom(  work~sons+income, data=womenlf,weight=ones )
mm5 <-multinom( work~sons+region,  data=womenlf,weight=ones )
mm5i <-multinom( work~sons*region,  data=womenlf,weight=ones )
mm6 <-multinom( work~sons*income,  data=womenlf,weight=ones )
mm7 <-multinom( work~sons+region+income,  data=womenlf,weight=ones )
mm8 <-multinom( work~region+sons*income, data=womenlf,weight=ones )

mm0$dev - mm1$dev
mm0$dev - mm3$dev
mm0$dev - mm4$dev

AIC(mm1,mm4,mm3) # mm3 434.8819
anova(mm1,mm3,test="Chisq")

summary(mm3)
#       log( xxxxtime/not_work):
#       logit 1: parttime vs not_work
#       logit 2: fulltime vs not_work

# residualPlots(mm3) # It does not work

summary(df)
mm3c <-multinom( work~ sons+I(income - mean(womenlf$income)), data=womenlf,weight=ones )
summary(mm3c)

exp(coef(mm3))
# Interpret sons-Absent and income on the mean
# logodd(PT/NW) = -1.33
# logodd(FT/NW) = 0.55
lopPT <- coef(mm3c)[1,1];lopPT
lopFT <- coef(mm3c)[2,1];lopFT
opPT <- exp(lopPT);opPT
opFT <- exp(lopFT);opFT
pNW <- (1/(1+opPT+opFT));pNW
pPT <- opPT * pNW; pPT
pFT <- opFT * pNW; pFT
pFT + pPT + pNW

predict(mm3c,type="class")
predict(mm3c, newdata=data.frame(sons="absent",income=14.76),type="probs")
predict(mm3c, newdata=data.frame(sons="absent",income=14.76),type="class")
```


```{r}
# Point 4
mmfit<- step(mm8) # mm3
anova(mm3,mm8,test='Chisq')
Anova(mm3, test="Chisq")
#residualPlots(mm3) # Not working
```


```{r}
# Point 6 and 7

anova(mm3,mm6,test='Chisq')
anova(mm1,mm3,test='Chisq')
anova(mm2,mm3,test='Chisq')
Anova(mm6,test="LR")
```


```{r}
# Point 8

mm33 <-multinom(  work~sons+income+I((income-mean(income))^2), data=womenlf,weight=ones )
anova(mm3,mm33,test='Chisq')

tt<-table(predict(mm3),womenlf$work);tt
100*sum(diag(tt))/sum(tt)
```


# Point 11

```{r}
mm3c <-multinom(  work~sons+I(income-mean(income)), data=womenlf,weight=ones )
summary(mm3c)
summary(mm3)

logodd2=0-1.3306078 + 0.02149908
logodd3=0.5480198 -2.55859890
logodd2;logodd3
odd2=exp(logodd2);odd2
odd3=exp(logodd3);odd3
prob1=1/(1+odd2+odd3)
prob2=prob1*odd2
prob3=prob1*odd3
prob1;prob2;prob3

predict(mm3c, type="probs")
predict(mm3c, type="probs",newdata=data.frame(sons="present",income=14.75665))

# Not working !
influence.measures(mm3)
cooks.distance(mm3)
sum(residuals(mm3,'deviance')^2)
sum(residuals(mm3,'pearson')^2)
```

## Hierarchical models

```{r}
# Point 12
########################################################################################
#       Hierarchical models:
#       Not Work    /   Work
#                       Fulltime / Parttime
#       Model Work / Not Work:

hm1m3 <- glm(bwork~income+sons,data=womenlf,family=binomial)
summary(hm1m3)

hm1m8 <- glm(bwork~income*sons + region,data=womenlf,family=binomial)
step(hm1m8)
```

# Point 13

```{r}
table( ifelse(predict(hm1m3,type="response")>0.5,"pwork","pnotwork"),womenlf$bwork)

library("ROCR")
dadesroc<-prediction(predict(hm1m3,type="response"),womenlf$bwork)
par(mfrow=c(1,2))
performance(dadesroc,"auc")
plot(performance(dadesroc,"err"))
plot(performance(dadesroc,"tpr","fpr"))
abline(0,1,lty=2)

# Interpret HL1 best model
coef(hm1m3)
exp(coef(hm1m3)[2:3])


# Prediction for a HH with sons 'present' and mean income
# Upper level

predict(hm1m3,type="response",newdata=data.frame(sons="present",income=14.75665))
1-predict(hm1m3,type="response",newdata=data.frame(sons="present",income=14.75665))
```

# Point 14

```{r}
summary(womenlf)
hm2m3 <- glm(work~income+sons,data=womenlf,family=binomial,subset=bwork=="work")
summary(hm2m3)
hm2m8 <- glm(work~income*sons + region,data=womenlf,family=binomial,subset=bwork=="work")
step( hm2m8 )

AIC(hm1m3)
AIC(hm2m3)
110.49+325.73
```

# Point 15 and 16

```{r}
# Lower level
# Diagnostics and prediction
residualPlots(hm2m3)
summary(subset(womenlf,work!="not_work"))
plot(allEffects(hm2m3))

#Coefficients: Full/Part
#                          Estimate Std. Error z value Pr(>|z|)    
# (Intercept)               1.89486    0.42476   4.461 8.16e-06 ***
# I(income - mean(income)) -0.10727    0.03915  -2.740  0.00615 ** 
# sonspresent              -2.65146    0.54108  -4.900 9.57e-07 ***
#Coefficients: Work/NotWork
#                          Estimate Std. Error z value Pr(>|z|)    
# (Intercept)               0.71150    0.24230   2.936  0.00332 ** 
# I(income - mean(income)) -0.04231    0.01978  -2.139  0.03244 *  
# sonspresent              -1.57565    0.29226  -5.391    7e-08 ***
#HL2
logodd3_2=1.89486-2.65146;logodd3_2
odd3_2=exp(logodd3_2);odd3_2
prob3cW=odd3_2/(1+odd3_2);prob3cW
prob2cW=1-prob3cW;prob2cW

predict(hm2m3,type="response",newdata=data.frame(sons="present",income=14.75665))
1-predict(hm2m3,type="response",newdata=data.frame(sons="present",income=14.75665))

#HL1
logoddW_1=0.71150-1.57565;logoddW_1
oddW_1=exp(logoddW_1);oddW_1
prW=oddW_1/(1+oddW_1);prW;
pr1=1-prW;pr1
pr2=prob2cW*prW;pr2
pr3=prob3cW*prW;pr3
pr1+pr2+pr3

table( ifelse(predict(hm1m3,type="response")>0.5,"pwork","pnotwork"),womenlf$bwork)
table(factor(ifelse(predict(hm2m3,type="response")<0.5,"pparttime","pfulltime"),levels=c("pparttime","pfulltime")),womenlf$work[womenlf$work!="not_work"])

100*(132+33+55)/263
```

# Point 17

```{r}

AIC(hm1m3)+AIC(hm2m3)
AIC(mm3) # Nominal proposal is better according to AIC. Nevertheless, confusion tgables

library(MNLpred)
# Plotting the predicted probabilities:
library(ggplot2)
library(scales)

# Hessian matrix is needed
womenlf$nsons <- as.numeric(womenlf$sons)-1
mm3 <- multinom(work~income+nsons,data=womenlf,Hess = TRUE)
summary(mm3)
pred1 <- mnl_pred_ova(model = mm3,
                      data = womenlf,
                      x = "income",
                      by = 1,
                      seed = "random", # default
                      nsim = 100, # faster
                      probs = c(0.025, 0.975)) # default

ggplot(data = pred1$plotdata, aes(x = income, 
                                  y = mean,
                                  ymin = lower, ymax = upper)) +
  geom_ribbon(alpha = 0.1) + # Confidence intervals
  geom_line() + # Mean
  facet_wrap(.~ work, scales = "free_y", ncol = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) + # % labels
  scale_x_continuous(breaks = c(0:10)) +
  theme_bw() +
  labs(y = "Predicted probabilities",
       x = "Income") # Always label your axes ;)

mm3 <- multinom(work~income+nsons,data=womenlf,Hess = TRUE)
summary(mm3)
fdif2 <- mnl_fd_ova(model = mm3,
                    data = womenlf,
                    x = "income",
                    by = 1,
                    z = "nsons",
                    z_values = c(0,1),
                    nsim = 100)

ggplot(data = fdif2$plotdata, aes(x = income, 
                                  y = mean,
                                  ymin = lower, ymax = upper,
                                  group = as.factor(nsons),
                                  linetype = as.factor(nsons))) +
  geom_ribbon(alpha = 0.1) +
  geom_line() +
  facet_wrap(. ~ work, scales = "free_y", ncol = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) + # % labels
  scale_x_continuous(breaks = c(0:10)) +
  scale_linetype_discrete(name = "Sons",
                          breaks = c(0, 1),
                          labels = c("Absent", "Present")) +
  theme_bw() +
  labs(y = "Predicted probabilities",
       x = "Income by Sons") # Always label your axes ;)
```

# Point 18: Ordinal models

```{r}
library(MASS)
om0 <-polr( work~1, data=womenlf,weight=ones )
om1 <-polr( work~sons, data=womenlf,weight=ones )
om2 <-polr( work~income, data=womenlf,weight=ones )
om4 <-polr( work~region, data=womenlf,weight=ones )
om3 <-polr(  work~sons+income, data=womenlf,weight=ones )
om5 <-polr( work~sons+region,  data=womenlf,weight=ones )
om5i <-polr( work~sons*region,  data=womenlf,weight=ones )
om6 <-polr( work~sons*income,  data=womenlf,weight=ones )
om7 <-polr( work~sons+region+income,  data=womenlf,weight=ones )
om8 <-polr( work~region+sons*income, data=womenlf,weight=ones )

anova(om0,om1,test='Chisq' )
anova(om0,om2,test='Chisq' )
anova(om0,om4,test='Chisq' )
Anova(om3,test='Chisq')
summary( om3 )
Anova( om8, test="LR")
step( om8 )

names(om8)
om0$deviance - om1$deviance
om0$deviance - om4$deviance
om0$deviance - om2$deviance
```

# Points 19 to 20

```{r}
anova(om5,om5i,test='Chisq')
anova(om1,om5,test='Chisq')

anova(om3,om6,test='Chisq')
anova(om1,om3,test='Chisq')
anova(om2,om3,test='Chisq')
```

# Punt 21

```{r}
om33 <-polr(  work~sons+income+I((income-mean(income))^2), data=womenlf,weight=ones )
anova(om3,om33,test='Chisq')
```

# Punt 22

```{r}

library(effects)
Effect(focal.predictors = c("income","sons"), om3)
plot(Effect(focal.predictors = c("income","sons"), om3), rug = FALSE)

summary(om3)

#Coefficients:
#                  Value Std. Error   t value
#income      -0.05390061 0.01949010 -2.765538
#sonspresent -1.97195668 0.28694823 -6.872169

#Intercepts:
#                  Value   Std. Error t value
#not_work|parttime -1.8520  0.3863    -4.7943
#parttime|fulltime -0.9409  0.3699    -2.5435

#Residual Deviance: 441.663
#AIC: 449.663

#       Explanatory variables appear with opposite sign as described in lecture notes
#       (LATENT VARIABLE).
# Interpret in probability terms: sons present and income on the mean 

om3c <-polr(  work~sons+I(income-mean(income)), data=womenlf,weight=ones )
summary(om3c)

logodd1=0-1.0566+1.97195668;logodd1
logodd2=0-0.1455+1.97195668;logodd2

gam1=exp(logodd1)/(1+exp(logodd1));gam1
gam2=exp(logodd2)/(1+exp(logodd2));gam2
pi1=gam1;pi1
pi2=gam2-gam1;pi2
pi3=1-gam2;pi3

predict(om3, type="probs", newdata=data.frame(sons="present",income=mean(womenlf$income)))
#       Most suitable model?
AIC(hm1m3)+AIC(hm2m3)
AIC(mm3) 
AIC(om3)# Nominal proposal is better according to AIC. Nevertheless, confusion tgables

# Confusion table
tt<-table(predict(om3),womenlf$work);tt
100*sum(diag(tt))/sum(tt)

  #          not_work parttime fulltime
  # not_work      133       36       23
  # parttime        0        0        0
  # fulltime       22        6       43

# Link
om3p <-polr(  work~sons+income, data=womenlf,weight=ones, method="probit" )
om3g <-polr(  work~sons+income, data=womenlf,weight=ones, method="cloglog" )
AIC( om3, om3p, om3g)  
summary( om3g )
```

# Housing

```{r}
options(contrasts=c("contr.treatment","contr.treatment"))
#options(contrasts = c("contr.treatment", "contr.poly"))

load("C:/Users/lmontero/Dropbox/DOCENCIA/FIB-SIM/2021-22/LABORATORY/Practice7/copenhaguen.RData")
summary(copen)

# Ordenar les categories

copen$housing <- ordered(copen$housing, levels=c("tower","apartments","atrium","terraced"))
copen$influence <- ordered(copen$influence, levels=c("low","medium","high"))
copen$contact <- ordered(copen$contact, levels=c("low","high"))
copen$satisfaction <- ordered(copen$satisfaction, levels=c("low","medium","high"))

summary(copen)
attach(copen)

#       Model de resposta nominal:
#       satisfaccio = housing + influence + contact

library(MASS)
library(nnet)
copen.mult <- multinom(satisfaction~housing+influence+contact,data=copen,weights=n)
summary(copen.mult)

#       Obtained estimates nominal outcome:
#       logit 1: medium vs low
#       logit 2: high vs low

# Coefficients:
#        (Intercept) housingapartments housingatrium housingterraced influencemedium
# medium  -0.4192316        -0.4356851     0.1313663      -0.6665728       0.4464003
# high    -0.1387453        -0.7356261    -0.4079808      -1.4123333       0.7348626
#        influencehigh contacthigh
# medium     0.6649367   0.3608513
# high       1.6126294   0.4818236

#       Ordinal Model :
#       satisfaction = housing * influence + contact

library(MASS)
copen.polr <- polr(satisfaction~housing*influence+contact,data=copen,weights=n)
summary(copen.polr)

copen.polrp <- polr(satisfaction~housing*influence+contact,data=copen,weights=n, method="probit" )
summary(copen.polrp)

copen.polrg <- polr(satisfaction~housing*influence+contact,data=copen,weights=n,  method="cloglog")
summary(copen.polrg)


#       Best treatment?
#       Ordinal response ?
#       Nominal response ?

#       Respondrem comparant les respostes observades i les esperades

# Això no funciona per ara: cal multiplicar pel tamany de cada classe.
tto<-xtabs(copen$n ~ copen$satisfaction+predict(copen.polr));tto
ttm<-xtabs(copen$n ~ copen$satisfaction+predict(copen.mult));ttm
100*sum(diag(tto))/sum(tto)
100*sum(diag(ttm))/sum(ttm)
# Take a look to AICs
# Ordinal: AIC: 3484.64 (logit)  AIC: 3485.331 (probit)  AIC: 3494.878 (cloglog)
# Multinomial: AIC: 3498.084  
AIC(copen.polr)
AIC(copen.polrp)
AIC(copen.polrg)
AIC(copen.mult)

# Visualització Nominal: wide structure
copen.wide <- expand.grid(housing=c("tower", "apartments", "atrium", "terraced"),
                     influence=c("low","medium", "high"),
                     contact=c("low", "high"))
copen.predm <- predict(copen.mult,copen.wide,type="probs"); copen.predm                    
copen.wide <- cbind( copen.wide, copen.predm )
library("reshape2")
copen.long <- melt (copen.wide,id.vars=c("housing","influence", "contact"),value.name="probablity")
copen.long
ggplot(copen.long, aes(x = housing, y = probablity, colour = influence)) +
       geom_point() + facet_grid(variable ~ ., scales="free")
ggplot(copen.long, aes(x = housing, y = probablity, colour = influence)) +
       geom_point() + facet_grid(variable ~ contact, scales="free")
ggplot(copen.long, aes(x = housing, y = probablity, colour = influence)) +
       geom_point(aes(x=housing,  size=contact)) + facet_grid(variable ~ .)

# Visualització Ordinal
library(effects)
Effect(focal.predictors = c("housing","influence","contact"), copen.polr)
plot(Effect(focal.predictors = c("housing","influence","contact"), copen.polr), rug = FALSE)


# Interpret zeta parameters
coef(copen.polr)
copen.polr$zeta
codds<-exp(copen.polr$zeta);codds
cprobs<-codds/(1+codds);cprobs
cprobset<-c(0,cprobs,1)
diff(cprobset)

# Interpret logit polr model for obs contact 'high', influence 'low' and housing 'tower'

coef(copen.polr)
eta<-copen.polr$zeta - coef(copen.polr)[6] # Pay attention to - sign
codds<-exp(eta);codds
cprobs<-codds/(1+codds);cprobs
cprobset<-c(0,cprobs,1)
diff(cprobset)
```

# FESAL 1985

Elaborated by German Rodríguez from the final report of the Demographic and Health Survey conducted in El Salvador in 1985 (FESAL-1985). Data table shows 3165 currently married women classified by age, grouped in five-year intervals, and current use of contraception, classified as sterilization, other methods, and no method.

```{r}
# Clear plots
if(!is.null(dev.list())) dev.off()

# Clean workspace
rm(list=ls())
options(contrasts=c("contr.treatment","contr.treatment"))
load("Fesal1985.RData")
summary(fesal85)
head(fesal85)
fesal85
```

## Nominal Proposal

```{r}
mm0 <- multinom( c.use ~ 1, weight=y, data=fesal85)
mm1 <- multinom( c.use ~ g.age, weight=y, data=fesal85)
mm2 <- multinom( c.use ~ n.age, weight=y, data=fesal85, Hess=T)
mm3 <- multinom( c.use ~ poly(n.age,2), weight=y, data=fesal85)
mm4 <- multinom( c.use ~ poly(n.age,3), weight=y, data=fesal85, Hess=T)
mm4m <- multinom( c.use ~ n.age + I(n.age^2) + I(n.age^3), weight=y, data=fesal85, Hess=T)
summary(mm1)
Anova( mm4m, test="LR")
summary(mm4m)
anova( mm0, mm1, test="Chisq")
anova( mm2, mm3, test="Chisq")
anova( mm3, mm4, test="Chisq")

AIC(mm1,mm4,mm2,mm3,mm4m)

# Visualization
# Observed data: probabilities and log-odds
ggplot(data = fesal85[,c(7,2,5)], aes(x = n.age, 
                                  y = oprob)) +
  geom_line() + # Mean
  # facet_wrap(.~ c.use, scales = "free_y", ncol = 3) +
    facet_wrap(.~ c.use, ncol = 3) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) + # % labels
  scale_x_continuous() +
  theme_bw() +
  labs(y = "Observed probabilities",
       x = "n.age") # Always label your axes ;)

ggplot(data = fesal85[,c(7,2,6)], aes(x = n.age, 
                                  y = ologodd)) +
  geom_line() + # Mean
  facet_wrap(.~ c.use, ncol = 3) + #, scales = "free_y"
  scale_y_continuous() + # % labels
  scale_x_continuous() +
  theme_bw() +
  labs(y = "Observed logodds",
       x = "n.age") # Always label your axes ;)

# Linear model in n.age predictions

pred1 <- mnl_pred_ova(model = mm2,
                      data = fesal85,
                      x = "n.age",
                      by = 5,
                      seed = "random", # default
                      nsim = 100, # faster
                      probs = c(0.025, 0.975)) # default

ggplot(data = pred1$plotdata, aes(x = n.age, 
                                  y = mean,
                                  ymin = lower, ymax = upper)) +
  geom_ribbon(alpha = 0.1) + # Confidence intervals
  geom_line() + # Mean
  facet_wrap(.~ c.use, ncol = 3) + #, scales = "free_y"
  scale_y_continuous(labels = percent_format(accuracy = 1)) + # % labels
  scale_x_continuous() +
  theme_bw() +
  labs(y = "Predicted probabilities",
       x = "n.age") # Always label your axes ;)

# Visualitzation considering factors
# Visualització Nominal: wide structure
fesal85.wide <- expand.grid(g.age=c("15-19", "20-24", "25-29", "30-34", "35-39", "40-44", "45-49"))
fesal85.predm <- predict(mm1,fesal85.wide,type="probs"); fesal85.predm                    
fesal85.wide <- cbind( fesal85.wide, fesal85.predm )
library("reshape2")
fesal85.long <- melt (fesal85.wide,id.vars=c("g.age"),value.name="probability")
fesal85.long <- cbind( fesal85.long , fesal85$oprob)
ggplot(fesal85.long, aes(x = g.age, y = probability, color=variable)) +
       geom_point() +  geom_line(aes( x=as.numeric(g.age), y=probability, color=variable ))       + facet_grid(variable ~ ., scales="free")+
  xlab("Factor Age")+ylab("Probability")+labs(title="Multinomial model mm1 - FESAL 85",color = "c.use") + geom_point( aes(x=g.age, y=fesal85$oprob), color = "black", size=2)

ggplot(fesal85.long, aes(x = g.age, y = probability, color=variable)) +
       geom_point() +  geom_line(aes( x=as.numeric(g.age), y=probability, color=variable )) +  xlab("Factor Age")+ylab("Probability")+labs(title="Multinomial model mm1 - FESAL 85",color = "c.use")

# Predict probabilities for c.use levels and someone having 22 years
summary(mm4m)
mm4m <- multinom( c.use ~ n.age + I(n.age^2) + I(n.age^3), weight=y, data=fesal85, Hess=T)
predict(mm4m,newdata=data.frame(n.age=22),type="probs")
predict(mm4m,newdata=data.frame(n.age=22),type="class")
loddOther<- coef(mm4m)["Other",]%*%c(1,22,22^2,22^3);loddOther
loddSter<- coef(mm4m)["Ster",]%*%c(1,22,22^2,22^3);loddSter
oddOther<-exp(loddOther)
oddSter<-exp(loddSter)
oddOther;oddSter

probNone <- 1/(1+oddOther+oddSter);probNone
probOther <- probNone*oddOther;probOther
probSter <- probNone*oddSter;probSter
probNone+probOther+probSter
```

## Ordinal Proposal

```{r}
library(MASS)
om0 <- polr( c.use ~ 1, weight=y, data=fesal85)
om1 <- polr( c.use ~ g.age, weight=y, data=fesal85)
om2 <- polr( c.use ~ n.age, weight=y, data=fesal85)
om3 <- polr( c.use ~ poly(n.age,2), weight=y, data=fesal85)
om4 <- polr( c.use ~ poly(n.age,3), weight=y, data=fesal85)
om3m <- polr( c.use ~ n.age + I((n.age)^2), weight=y, data=fesal85)
om3mp <- polr( c.use ~ n.age + I((n.age)^2), weight=y, data=fesal85, method="probit")
om3mg <- polr( c.use ~ n.age + I((n.age)^2), weight=y, data=fesal85, method="cloglog")

summary(om3m)
summary(om3mp)
Anova(om4,test="Chisq")

anova( om1, om4, test="Chisq" ) # Not correct
anova( om2, om3, test="Chisq" )
anova( om3, om4, test="Chisq" )

# Best model Y ~ poly( X,2 )

AIC( mm4, om3m, om1, om3mp, om3mg )

# Visualització Ordinal
library(effects)
plot(Effect(focal.predictors = c("n.age"), om3m))

# Forecast probability of categories for a 22 year woman

om3mc <- polr( c.use ~ n.age + I(n.age^2), weight=y, data=fesal85)
predict(om3mc,newdata=data.frame(n.age=22),type="probs")
predict(om3mc,newdata=data.frame(n.age=22),type="class")

coef(om3mc)
om3mc$zeta
lcoddNone<- om3mc$zeta[1] - coef(om3mc)%*%c(22,22^2);lcoddNone
lcoddOther<- om3mc$zeta[2] - coef(om3mc)%*%c(22,22^2);lcoddOther
coddNone<-exp(lcoddNone)
coddOther<-exp(lcoddOther)
coddNone;coddOther

cprobNone <- coddNone/(1+coddNone);cprobNone
cprobOther <- coddOther/(1+coddOther);cprobOther
probSter <- 1-cprobOther;probSter
cprobNone;cprobOther-cprobNone;probSter
```

## Hierarchical Approach

```{r}
fesal85$bc.use <- factor(ifelse(fesal85$c.use=="None",0,1), labels = c( "None", "Some" ) )

bh1m0 <- glm( bc.use ~ 1, family = binomial, data = fesal85, weights = y )
bh1m1 <- glm( bc.use ~ n.age, family = binomial, data = fesal85, weights = y )
bh1m2 <- glm( bc.use ~ poly(n.age,2), family = binomial, data = fesal85, weights = y )
bh1m3 <- glm( bc.use ~ poly(n.age,3), family = binomial, data = fesal85, weights = y )

Anova( bh1m2, test="LR" )
anova( bh1m1, bh1m2, test="Chisq" )
anova( bh1m2, bh1m3, test="Chisq" )

library(effects)
plot(allEffects( bh1m2 ))

bh2m0 <- glm( c.use ~ 1, family = binomial, data = fesal85[fesal85$bc.use=="Some",], weights = y )
bh2m1 <- glm( c.use ~ n.age, family = binomial, data = fesal85[fesal85$bc.use=="Some",], weights = y )
bh2m2 <- glm( c.use ~ poly(n.age,2), family = binomial, data = fesal85[fesal85$bc.use=="Some",], weights = y )
bh2m3 <- glm( c.use ~ poly(n.age,3), family = binomial, data = fesal85[fesal85$bc.use=="Some",], weights = y )

Anova( bh2m2, test="LR" )
anova( bh2m1, bh2m2, test="Chisq" )
anova( bh2m2, bh2m3, test="Chisq" )

plot(allEffects( bh1m2 ))
plot(allEffects( bh2m3 ))


# Prediction 22 years
bh1 <- glm( bc.use ~ I((n.age-22)) + I((n.age-22)^2), family = binomial, data = fesal85, weights = y )
bh2 <- glm( c.use ~ I((n.age-22)) + I((n.age-22)^2) + I((n.age-22)^3), family = binomial, data = fesal85[fesal85$bc.use=="Some",], weights = y )

coef(bh1)
coef(bh2)
exp(coef(bh1))[1]
exp(coef(bh2))[1]
probSome <- exp(coef(bh1))[1]/(1+exp(coef(bh1))[1]);probSome
probOthercSome <- 1-(exp(coef(bh2))[1]/(1+exp(coef(bh2))[1]));probOthercSome
probStercSome <-(exp(coef(bh2))[1]/(1+exp(coef(bh2))[1]));probStercSome
probNone <-1-probSome;probNone
probOther <- probSome*probOthercSome;probOther
probSter <- probSome*probStercSome;probSter
probNone+probOther+probSter
```


```{r}
# AIC
# 
AIC(bh1m2) + AIC(bh2m3)
AIC(mm4)
AIC(om3m)

# Confusion tables

ttm <- xtabs( y ~ c.use + predict( mm4 ), data=fesal85); ttm
100*sum(diag(ttm))/sum(ttm)
tto <- xtabs( y ~ c.use + predict( om3m ), data=fesal85); tto
100*sum(diag(tto))/sum(tto)

ttb1 <- xtabs( y ~ bc.use + ifelse(predict( bh1m2, type = "response" )< 0.5, 0, 1), data=fesal85); ttb1
ttb2 <- xtabs( y ~ c.use + ifelse(predict( bh2m3, newdata= fesal85[fesal85$bc.use=="Some",], type = "response" )< 0.5, 0, 1), data=fesal85[fesal85$bc.use=="Some",]); ttb2
100*sum(815+198+922)/sum(ttm)
```

# Cheese tasting

```{r}
load("C:/Users/lmontero/Dropbox/DOCENCIA/FIB-SIM/2021-22/LABORATORY/Practice7/cheese.RData")
summary(cheese)
```

```{r}
om0 <- polr( c.taste ~ 1, data = cheese, weights=y )
om1 <- polr( c.taste ~ additive, data = cheese, weights=y )
om1p <- polr( c.taste ~ additive, data = cheese, method="probit",weights=y )
om1cg <- polr( c.taste ~ additive, data = cheese, method="cloglog",weights=y )
anova( om0 , om1, test="Chisq")
summary(om1)
summary(om1p)
summary(om1cg)

AIC( om1, om1p, om1cg )

om1p$zeta
cpA <- c(0,pnorm(om1p$zeta),1); cpA
pA<-diff(cpA);pA

om1p$coef
om1p$zeta - om1p$coef[3]
cpD <- c(0,pnorm(om1p$zeta - om1p$coef[3]),1); cpD
pD<-diff(cpD);pD

# Visualització Ordinal
library(effects)
cex=0.5
plot(Effect(focal.predictors = c("additive"), om1,cex=0.5))

```