---
title: "BSG-MDS practical 5 Statistical Genetics"
author: "Eliya Tiram and Ximena Moure"
date: "05/12/2023, submission deadline 12/12/2023"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
#install.packages("genetics")
library(genetics)
```


## Setting up data
```{r}
cases <- c(112, 278, 150)
controls <- c(206, 348, 150)
```


## 1. Perform the alleles test for this data set. Provide the p-value and the odds ratio and comment on the results.

We get a p-value of 0.0002100232 which is very small. It is well below the 
conventional threshold of 0.05 for statistical significance. This indicates that
the difference in allele frequencies between the cases and controls in the 
study is highly unlikely to have occurred by chance. 
There is a statistically significant association between the SNP and 
Alzheimer's disease.

Since the odds ratio is less than 1, it suggests that allele A is associated 
with a decreased risk of Alzheimer's disease compared to allele a. However, it's
important to note that the odds ratio is less than 1, which implies a protective
effect rather than a risk factor. In other words, individuals with allele A may
be less likely to develop Alzheimer's disease compared to those with allele a.

```{r}
allele_A_cases <- 2 * cases[1] + cases[2]
allele_a_cases <- 2 * cases[3] + cases[2]
allele_A_controls <- 2 * controls[1] + controls[2]
allele_a_controls <- 2 * controls[3] + controls[2]


mat <- matrix(c(allele_A_cases, allele_A_controls, allele_a_cases, allele_a_controls), nrow = 2)
colnames(mat) <- c("A","a")
rownames(mat) <- c("Cases","Controls")

# Chi-square test for allele frequencies
chi_test <- chisq.test(mat,correct=FALSE)

fi <- fisher.test(mat)

# Output
cat("P-value:", chi_test$p.value, "\n")
cat("Odds ratio: ", fi$estimate, "\n")
```

## 2. Test for association using a codominant, a dominant and a recessive model. Provide the p-values for all the tests and comment on the results.

All three models show statistical significance, which suggests that the genetic 
variant is associated with the disease risk. However, the codominant and 
dominant models have lower p-values, indicating stronger evidence for these 
models compared to the recessive model.

Given that the p-values for codominant and dominant models are very close, 
it might suggest that the risk increases with one copy of the 'm' allele and 
potentially further increases (although not dramatically) with a second copy. 
This could point towards a codominant pattern where each allele adds to the 
risk, but the presence of one allele is sufficient for an increased risk, 
which might also be consistent with a partial dominant effect.

```{r}
gen_count <- matrix(c(112, 278, 150, 206, 348, 150), nrow = 2, byrow = TRUE)
colnames(gen_count) <- c("AA","Aa","aa")
rownames(gen_count) <- c("Cases","Controls")

# Codominant Test
test_codominant <- chisq.test(gen_count)
p_value_codominant <- test_codominant$p.value
fisher_codominant <- fisher.test(gen_count)
cat("Codominant Test Results:\n")
cat("P-value Pearson:", p_value_codominant, "\n")
cat("P-value Fisher: ", fisher_codominant$p.value, "\n")

# Dominant Test
Y_dominant <- cbind(gen_count[,1], gen_count[,2] + gen_count[,3])
test_dominant <- chisq.test(Y_dominant)
p_value_dominant <- test_dominant$p.value
fisher_dominant <- fisher.test(Y_dominant)
cat("\nDominant Test Results:\n")
cat("P-value Pearson:", p_value_dominant, "\n")
cat("P-value Fisher: ", fisher_dominant$p.value, "\n")

# Recessive Test
Y_recessive <- cbind(gen_count[,1] + gen_count[,2], gen_count[,3])
test_recessive <- chisq.test(Y_recessive)
p_value_recessive <- test_recessive$p.value
fisher_recessive <- fisher.test(Y_recessive)
cat("\nRecessive Test Results:\n")
cat("P-value Pearson:", p_value_recessive, "\n")
cat("P-value Fisher: ", fisher_recessive$p.value, "\n")

```

## 3. Plot the risk of disease as a function of the number of m alleles. Comment on the results. Which model seems most appropriate?

Since the risk of disease increases with the number of m alleles, the additive
model seems more appropriate.

The smooth increase in risk from 'AA' to 'aa' suggests that neither a purely 
dominant nor a purely recessive model may fully explain the genetic influence 
on disease risk.

```{r}
risk_AA <- cases[1] / (cases[1] + controls[1])
risk_Aa <- cases[2] / (cases[2] + controls[2])
risk_aa <- cases[3] / (cases[3] + controls[3])
risks <- c(risk_AA, risk_Aa, risk_aa)

labels <- c("AA", "Aa", "aa")
x_pos <- 1:length(labels)

plot(x_pos, risks, type = "b", col = "blue", xlab = "Number of m alleles",
     ylab = "Risk of Disease", pch = 19, xaxt = "n" )
title("Risk of Disease by Number of m Alleles")
axis(1, at = x_pos, labels = labels)
```

## 4. Perform Armitage trend test for this data set. Does the null hypothesis β1 = 0 hold? Comment on your response.

Given the p-value is less than the typical threshold for statistical 
significance (0.05), we reject the null hypothesis. This suggests that there is 
a statistically significant association between the genetic variant and 
Alzheimer's disease.

```{r}
X <- rbind(cases,controls)
n <- sum(X)
cas <- rep(c(0,1,2),cases)
con <- rep(c(0,1,2),controls)
y <- c(rep(1,sum(cases)), rep(0,sum(controls)))
x <- c(cas,con)
r <- cor(x,y)
A <- n*(r^2)
pvalue <- pchisq(A,df=1,lower.tail=FALSE)
cat("Armitage trend test:", A,"\n")
cat("P-value:", pvalue, "\n")
```



## 5. Is there evidence for association of this marker with the disease? Argument your response.

The plots for the codominant, dominant, and recessive models all show an 
increase in risk with the presence of the 'm' allele(s), suggesting a 
relationship between the allele and disease risk.
The p-values for all three genetic models are below the standard threshold of 
0.05, indicating that the results are statistically significant and the 
associations observed are unlikely due to chance.
The p-value from the Armitage trend test, which is particularly low 
(0.0002000008), provides strong evidence against the null hypothesis of no 
association.
Together, these elements point to a consistent pattern of association between 
the genetic marker in question and the disease. 
The evidence suggests that this marker is likely to be related to the risk of 
developing the disease.



