---
title: "BSG-MDS practical 4 Statistical Genetics"
author: "Eliya Tiram and Ximena Moure"
date: "28/11/2023, submission deadline 05/12/2023"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, echo=FALSE, message=FALSE, warning=FALSE}
#install.packages("haplo.stats")
library(dplyr)
library(genetics)
library(MASS)
library(data.table)

```

## Load data
```{r}
data <- fread("Chr21.dat")
genetic_data <- data[, -c(1:6), with=FALSE]
```

## 1. How many variants are there in this database? What percentage of the data is missing?
```{r}
num_variants <- ncol(genetic_data)
missing_data_percentage <- 100 * sum(is.na(genetic_data)) / (nrow(genetic_data) * ncol(genetic_data))

cat("\nNum variants:",num_variants, "\n")
cat("\nPercentage missing:",missing_data_percentage, "\n")

```

## 2. Compute the Manhattan distance matrix between the individuals (which is identical to the Minkowsky distance with parameter λ = 1) using R function dist. Include a submatrix of dimension 5 by 5 with the distances between the first 5 individuals in your report.

```{r}
manhattan_distances <- as.matrix(dist(genetic_data, method = "manhattan"))
# Extract a 5x5 submatrix for the first five individuals
submatrix_5x5 <- manhattan_distances[1:5, 1:5]

submatrix_5x5
```

## 3. How does the Manhattan distance relate to the allele sharing distance?

Manhattan distance calculates the number of different alleles between two 
polymorphisms, and the allele sharing distance does the same from the encoded 
version (0,1,2) by extracting the difference.

```{r}


```


## 4. Apply metric multidimensional scaling (cmdscale) with two dimensions, k = 2, using the Manhattan distance matrix and include the map in your report. Do you think the data come from one homogeneous human population? If not, how many subpopulations do you think the data might come from, and how many individuals pertain to each suppopulation?

In the plot, we see two distinct clusters of points. This suggests that the 
individuals in this dataset likely come from at least two different 
subpopulations. 
The clusters are quite distinct and do not overlap, indicating that the genetic
variation between the groups is substantial.

```{r}
mds <- cmdscale(manhattan_distances, k = 2, eig = T)
mds_scaled <- as.data.frame(mds$points)
# Plot the map
plot(mds_scaled$V1,mds_scaled$V2, main = "MDS Plot", xlab = "Dimension 1", ylab = "Dimension 2")

```

```{r}
threshold <- 0
num_subpop1 <- sum(mds$points[,1] < threshold)
num_subpop2 <- sum(mds$points[,1] >= threshold)
num_subpop1
num_subpop2

```

We could also use kmeans to calculate the number of individuals in each 
subpopulation.

```{r}
set.seed(42)
kmeans_result <- kmeans(mds$points, centers = 2)

# Count the number of individuals in each cluster
table(kmeans_result$cluster)

plot(mds$points[, 1], mds$points[, 2], col = kmeans_result$cluster,
     xlab = "Dimension 1", ylab = "Dimension 2", main = "MDS Plot with K-Means Clustering")
legend("topright", legend = c("Subpopulation 1", "Subpopulation 2"), 
       col = 1:2, pch = 1)


```
## 5. What is the goodness-of-fit of the two-dimensional approximation to your distance matrix? Explain which criterium you have used.

The goodness-of-fit measure (GOF) that we obtain from the mds$GOF is based on 
the proportion of variance accounted for by the chosen number of dimensions.

First Value (0.1703581): This represents the variance accounted for by the first
dimension alone. It suggests that the first dimension captures about 17.03% of 
the total variance.
Second Value (0.1703605): This represents the cumulative variance accounted for 
by both the first and second dimensions. Since the increase from the first to 
the second value is very small, it implies that the second dimension adds very 
little additional information beyond what is captured by the first dimension.

A cumulative goodness-of-fit of 17.04% is relatively low. This means that the 
two-dimensional solution does not capture a large portion of the variance in 
the original high-dimensional distance data.

```{r}
gof <-mds$GOF
cat("\nGOF:",gof, "\n")
```


## 6. Make a plot of the estimated distances (according to your two-dimensional map of individuals) versus the observed distances. What do you observe? Regress estimated distances on observed distances and report the coefficient of determination of the regression (you can use the function lm).

```{r}
mds_distances <- as.matrix(dist(mds$points))

# Plot the estimated MDS distances against the observed Manhattan distances
plot( mds_distances, manhattan_distances, main = "Estimated vs. Observed Distances",
     xlab = "Estimated Distances (MDS)", ylab = "Observed Distances (Manhattan)")

# Perform a linear regression of observed distances on estimated MDS distances
regression <- lm(manhattan_distances ~ mds_distances)

# Extract the R-squared value from the regression summary
r_squared <- summary(regression)$r.squared

cat("\nCoefficient of Determination (R-squared):", r_squared, "\n")
```
```{r}

mds_distances <- as.matrix(dist(mds$points))


observed_distances <- as.vector(as.matrix(manhattan_distances))
estimated_distances <- as.vector(mds_distances)


plot(observed_distances, estimated_distances, xlab = "Observed Distances", ylab = "Estimated Distances", main = "Observed vs Estimated Distances")
abline(0, 1, col = "red", lty = 1)

fit <- lm(estimated_distances ~ observed_distances)
summary_fit <- summary(fit)
summary(fit)

r_squared <- summary_fit$r.squared
print(paste("Coefficient of Determination (R-squared):", r_squared))
```

## 7. We now try a (two-dimensional) non-metric multidimensional scaling using the isoMDs function that you will find in MASS library. We use a random initial configuration and, for the sake of reproducibility, make this random initial configuration with the instructions: set.seed(12345) and init <- scale(matrix(runif(m*n),ncol=m),scale=FALSE) where n represents the sample size and m represents the dimensionality of the solution. Make a plot of the two-dimensional solution. Do the results support that the data come from one homogeneous population?

There are no obvious subgroups indicating separate populations, this can suggest 
that the individuals in the dataset might come from a single homogeneous 
population.

```{r}
set.seed(12345)
n <- nrow(genetic_data)
m <- 2
init <- scale(matrix(runif(m * n), ncol = m), scale = FALSE)
nonmetric_mds <- isoMDS(manhattan_distances, k = 2, y = init)
# Plot the non-metric MDS map
plot(nonmetric_mds$points, main = "Non-Metric MDS Plot", xlab = "Dimension 1", ylab = "Dimension 2")

```

## 8. Try some additional runs of the two-dimensional isoMDS with different initial configurations. Make a plot of the solutions and report the STRESS for each of them. What do you observe?

The variation in stress values implies that the initial configuration has a 
significant impact on the resulting MDS solution. The results suggest that the 
optimization process in isoMDS might be getting stuck in local minima, leading 
to suboptimal solutions in some runs.

From the plots we can observe that some of them show 1 population and others 2.

```{r}
library(MASS)
library(ggplot2)

times <- 8
par(mfrow = c(2, 3))
stress_r <- numeric(times) # To store stress values
mds_run_results <- list() # To store MDS results for each run

set.seed(123)
for (i in 1:times) {
  # Random initial configuration
  init <- scale(matrix(runif(2 * nrow(manhattan_distances)), ncol = 2), scale = FALSE)
  mds_run_results[[i]] <- isoMDS(manhattan_distances, k = 2, y = init)$points
  plot(mds_run_results[[i]], main = paste("Run", i), xlab = "D1", ylab = "D2", pch = 19, col = "red")
  stress_r[i] <- isoMDS(manhattan_distances, k = 2, y = init)$stress
}

par(mfrow = c(1, 1))
```

```{r}
cat("Stress values: ", stress_r, "\n")
```

## 9.Compute the stress for a 1, 2, 3, . . . , 50-dimensional solution. How many dimensions are necessary to obtain a good representation with a stress below 10? Make a plot of the stress against the number of dimensions.

```{r}

stress_values <- numeric(50)  # to store stress values for each dimension
dimensions <- 1:50  # vector containing the number of dimensions
cmdscale_result <- cmdscale(manhattan_distances, eig = TRUE, k = max(dimensions))

for (k in dimensions) {
   init <- cmdscale_result$points[, 1:k, drop = FALSE]
  # Run non-metric MDS with varying dimensions
  mds_result <- isoMDS(manhattan_distances, k = k, y=init)
  # Store the stress value
  stress_values[k] <- mds_result$stress
}

# Find the number of dimensions needed for stress below 10
good_dims <- which(stress_values < 10)

if (length(good_dims) > 0) {
  cat("Minimum dimensions needed for stress below 10:", min(good_dims), "\n")
} else {
  cat("No solution with stress below 10 found up to 50 dimensions.\n")
}

# Plot stress values against number of dimensions
plot(dimensions, stress_values, type = "b", xlab = "Number of Dimensions", ylab = "Stress Value", main = "Stress vs. Number of Dimensions")

```

## 10. Run the two-dimensional isoMDS a hundred times, each time using a different random initial configuration using the instructions above. Report the stress of the best and the worse run, and plot the corresponding maps. Compare your results to the metric MDS and comment on your findings.

The plot for the best run shows a more defined structure with data points 
grouped together, indicating that the two-dimensional space may be 
capturing some meaningful relationships within the data.

The plot for the worst run shows the points dispersed in a circular pattern, 
which suggests that the two-dimensional representation is more or less random 
and does not capture the structure of the data effectively.

The substantial difference in stress values and the visual configurations 
imply that the initial configuration can significantly influence the quality of 
the isoMDS solution.




```{r}


n_runs <- 100
n <- nrow(manhattan_distances)
stress_values <- numeric(n_runs)
mds_configs <- list()

for (i in 1:n_runs) {
  set.seed(i)
  init <- scale(matrix(runif(2 * n), ncol = 2), scale = FALSE) # Random initial configuration
  mds_run <- isoMDS(manhattan_distances, k = 2, y = init)
  stress_values[i] <- mds_run$stress
  mds_configs[[i]] <- mds_run$points
}

# Identify the best and worst runs
best_run <- which.min(stress_values)
worst_run <- which.max(stress_values)

cat("Best run stress:", stress_values[best_run], "\n")
cat("Worst run stress:", stress_values[worst_run], "\n")

# Plot the best and worst MDS configurations
par(mfrow=c(1, 2)) # Set up plotting area

plot(mds_configs[[best_run]][,1], mds_configs[[best_run]][,2], 
     main = paste("Best Run (Stress:", round(stress_values[best_run], 4), ")"),
     xlab = "Dimension 1", ylab = "Dimension 2", pch = 19)

plot(mds_configs[[worst_run]][,1], mds_configs[[worst_run]][,2], 
     main = paste("Worst Run (Stress:", round(stress_values[worst_run], 4), ")"),
     xlab = "Dimension 1", ylab = "Dimension 2", pch = 19)

par(mfrow=c(1, 1))

```


## 11. Compute the correlation matrix between the first two dimensions of the metric MDS and the two-dimensional solution of your best non-metric MDS. Comment your findings.

The first dimension of metric MDS (Dimension 1) has a very high positive 
correlation with the first dimension of non-metric MDS. This indicates that 
both MDS methods are in strong agreement and are capturing a similar pattern of
variation along their respective first dimensions.

The first dimension of metric MDS also has a high negative correlation with the
second dimension of non-metric MDS. A negative correlation means that as one 
dimension increases, the other decreases. 

There is a low negative correlation between the second dimension of metric MDS 
and both dimensions of non-metric MDS. These low values suggest that the second 
dimension of the metric MDS does not correspond well to either dimension of the 
non-metric MDS, indicating dissimilarity in the patterns captured by this 
particular dimension across the two methods.

The strong correlations (positive and negative) indicate that while there is 
some agreement between the methods on the primary dimension of variation, there 
is a significant difference in how the secondary dimensions are represented.



```{r}

# best non-metric MDS run
best_nonmetric_mds_points <- mds_configs[[best_run]]

cor_matrix <- cor(mds_scaled, best_nonmetric_mds_points)

print(cor_matrix)

image(1:2, 1:2, as.matrix(cor_matrix), main="Correlation Matrix", xlab="Metric MDS Dimensions", ylab="Non-metric MDS Dimensions", xaxt='n', yaxt='n')
axis(1, at=1:2, labels=c("Dimension 1", "Dimension 2"))
axis(2, at=1:2, labels=c("Dimension 1", "Dimension 2"), las=2)  # 'las=2' makes the y-axis labels perpendicular to the axis

```